---
title: "Using commonMZ rules with CAMERA"
subtitle: "Annotating adducts and fragments in an XCMS peak table"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 5
    theme: cosmo
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Using commonMZ rules with CAMERA}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(dplyr)
library(DT)
```

CAMERA's `findAdducts()` function needs a table of adduct and fragment rules: one
row per rule, with columns `name`, `nmol`, `charge`, `massdiff`, `oidscore`,
`quasi`, and `ips`. `MZ_CAMERA()` builds exactly that table from commonMZ's curated
`adducts_fragments` dataset. This vignette shows what the table looks like, how to
customise it, and where it plugs into a CAMERA workflow.

## Building the rules table

`MZ_CAMERA()` accepts `"pos"`, `"neg"`, or `"both"` (which returns positive and
negative rules combined). A `warn_clash` argument flags pairs of rules whose mass
differences are indistinguishable within a given ppm tolerance:

```{r rules-pos}
rules_pos <- MZ_CAMERA(mode = "pos", warn_clash = TRUE, clash_ppm = 5)
nrow(rules_pos)
```

Browse the full table: sort any column or use the search box to filter by name or
mass difference.

```{r rules-pos-table}
#| echo: false
rules_pos %>%
  mutate(massdiff = round(massdiff, 4)) %>%
  datatable(rownames = FALSE,
            options = list(pageLength = 15, order = list(list(3, "asc"))))
```

## The NH₄⁺ clash

A common clash in positive mode: `[M+NH4]+` and neutral loss of NH₃ produce the
same 17.027 Da difference and cannot be distinguished on mass alone. The warning
from `warn_clash = TRUE` flags this. If NH₄⁺ adducts are rare in your matrix
(common for many LC-MS setups), remove them before passing the rules to CAMERA:

```{r remove-nh4}
rules_pos_clean <- rules_pos %>% filter(name != "[M+NH4]+")
nrow(rules_pos_clean)
```

## Negative mode

```{r rules-neg}
rules_neg <- MZ_CAMERA(mode = "neg", warn_clash = TRUE, clash_ppm = 5)
```

```{r rules-neg-table}
#| echo: false
rules_neg %>%
  mutate(massdiff = round(massdiff, 4)) %>%
  datatable(rownames = FALSE,
            options = list(pageLength = 15, order = list(list(3, "asc"))))
```

## Plugging into CAMERA

Pass the rules table as the `rules` argument to `findAdducts()`. CAMERA requires a
plain `data.frame`, not a tibble, so convert with `as.data.frame()`:

```{r camera-sketch}
#| eval: false
library(CAMERA)

# after xsAnnotate() and groupFWHM() / groupCorr() / findIsotopes():
cam_pos <- findAdducts(cam_pos_iso,
                       ppm       = 10,
                       mzabs     = 0.01,
                       multiplier = 4,
                       polarity  = "positive",
                       rules     = as.data.frame(rules_pos_clean))

cam_neg <- findAdducts(cam_neg_iso,
                       ppm       = 10,
                       mzabs     = 0.01,
                       multiplier = 4,
                       polarity  = "negative",
                       rules     = as.data.frame(rules_neg))
```

`multiplier` controls the maximum number of molecules per ion cluster CAMERA will
consider; 3 or 4 covers most practical cases. `ppm` and `mzabs` set the matching
tolerance: CAMERA applies whichever window is wider, so for high-resolution data
set `mzabs` to something small (0.005 or 0.01) and rely on `ppm`.

## Checking annotation frequency

After running `findAdducts()`, it is useful to see which rules actually fired.
Rules that appear zero times are either absent from your sample or the tolerance
was too tight:

```{r freq-sketch}
#| eval: false
rules_count <- cam_pos@annoID[, "ruleID"] %>%
  factor(levels = seq_len(nrow(rules_pos_clean))) %>%
  table() %>%
  as.data.frame() %>%
  setNames(c("ruleID", "n")) %>%
  mutate(name = rules_pos_clean$name[as.integer(ruleID)]) %>%
  arrange(desc(n))

head(rules_count, 15)
```

If you want to browse the mass differences in this table against a broader reference
(including repeating units), see
[*Looking up an unexplained mass difference*](mass-difference-lookup.html).
