---
title: "Using commonMZ rules with CAMERA"
subtitle: "Annotating adducts and fragments in an LC-MS peak table"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 5
    theme: cosmo
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Using commonMZ rules with CAMERA}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(dplyr)
```

```{r fix-dt-core}
#| echo: false
# pkgdown strips jquery.dataTables.min.js from widget deps (jQuery dedup false-positive).
# Inline the DataTables plugin so the DT tables render correctly.
htmltools::includeScript(
  system.file("htmlwidgets/lib/datatables/js/jquery.dataTables.min.js", package = "DT")
)
```

CAMERA's `findAdducts()` function needs a table of adduct and fragment rules: one row per rule, with columns `name`, `nmol`, `charge`, `massdiff`, `oidscore`, `quasi`, and `ips`. `MZ_CAMERA()` builds exactly that table from commonMZ's curated `adducts_fragments` dataset.

The seven columns are defined as follows. `name`, `nmol`, `charge`, and `massdiff` are described in Kuhl et al. (2012); `oidscore`, `quasi`, and `ips` are documented only in the CAMERA source code.

| Column | Type | Meaning |
|------------------------|------------------------|------------------------|
| `name` | character | Ion label (e.g. `[M+H]+`). |
| `nmol` | integer | Molecules per ion (1 = monomer, 2 = dimer). |
| `charge` | integer | Signed ion charge. |
| `massdiff` | numeric (Da) | Mass added by the adduct (compared to neutral mass) or lost as a neutral fragment. |
| `oidscore` | integer | Groups rules that share the same adduct formula but differ in `nmol` (e.g. `[M+H]+` and `[2M+H]+`). CAMERA uses it to automatically extend a monomer match to the corresponding dimer/trimer hypotheses. |
| `quasi` | 0 / 1 | **Mandatory ion flag.** If no rule with `quasi = 1` appears in a candidate group, CAMERA discards the whole group. Typically only the primary protonated ion carries `quasi = 1`. |
| `ips` | numeric | **Ion/adduct peak score** (\~0.25–1). Confidence weight per rule. When a peak fits multiple competing mass hypotheses, CAMERA sums `ips` within each group and discards the lower-scoring one. Typical values: 1.0 for `[M+H]+`, 0.5 for multiply-charged ions, 0.25 for neutral losses. |

## Building the rules table

`MZ_CAMERA()` accepts three modes: `"pos"` for positive-ion LC-MS, `"neg"` for negative-ion LC-MS, and `"ei"` for electron-ionisation GC-MS. The `warn_clash` argument flags pairs of rules whose mass differences fall within a given ppm tolerance of each other and cannot be distinguished on mass alone.

### Positive mode

```{r rules-pos}
rules_pos <- MZ_CAMERA(mode = "pos", warn_clash = TRUE, clash_ppm = 5)
```

Browse the full table — sort any column or use the search box to filter by name or mass difference:

```{r rules-pos-table}
#| eval: false
rules_pos %>% mutate(massdiff = round(massdiff, 4))
```

```{r rules-pos-table-dt}
#| echo: false
rules_pos %>% mutate(massdiff = round(massdiff, 4)) %>%
  DT::datatable(rownames = FALSE,
                options = list(pageLength = 15, order = list(list(3, "asc"))))
```

### Negative mode

```{r rules-neg}
rules_neg <- MZ_CAMERA(mode = "neg", warn_clash = TRUE, clash_ppm = 5)
nrow(rules_neg)
```

```{r rules-neg-table}
#| code-fold: true
#| eval: false
rules_neg %>% mutate(massdiff = round(massdiff, 4))
```

```{r rules-neg-table-dt}
#| echo: false
rules_neg %>% mutate(massdiff = round(massdiff, 4)) %>%
  DT::datatable(rownames = FALSE,
                options = list(pageLength = 15, order = list(list(3, "asc"))))
```

### Electron ionisation (EI)

```{r rules-ei}
#| code-fold: true
rules_ei <- MZ_CAMERA(mode = "ei", warn_clash = FALSE)
```

## The NH₄⁺ clash

A common clash in positive mode: `[M+NH4]+` and neutral loss of NH₃ produce the same 17.027 Da difference and cannot be distinguished on mass alone. The warning from `warn_clash = TRUE` flags this. If NH₄⁺ adducts are rare in your matrix (common for many reversed-phase LC-MS setups), remove them before passing the rules to CAMERA:

```{r remove-nh4}
rules_pos_clean <- rules_pos %>% filter(name != "[M+NH4]+")
```

## Complete example: CAMERA's built-in data

The steps below annotate `mm14`, CAMERA's built-in example xcmsSet (134 peaks, single-sample, positive mode), using commonMZ rules. The workflow has five stages:

**Step 1 — create an `xsAnnotate` object** from an xcms result:

```{r step1}
#| eval: false
library(CAMERA)
library(xcms)

# xs is an xcmsSet produced by xcms peak detection
cam_pos <- xsAnnotate(xs, polarity = "positive")
```

**Step 2 — group co-eluting peaks** by retention time window (FWHM-based):

```{r step2}
#| eval: false
cam_pos <- groupFWHM(cam_pos, perfwhm = 0.6, intval = "into")
```

**Step 3 (optional) — refine groups by cross-sample correlation:**

```{r step3}
#| eval: false
cam_pos <- groupCorr(cam_pos, cor_eic_th = 0.7, pval = 1e-6,
                     calcIso = FALSE, calcCiS = FALSE, calcCaS = TRUE)
```

**Step 4 — annotate isotope peaks:**

```{r step4}
#| eval: false
cam_pos <- findIsotopes(cam_pos, ppm = 10, mzabs = 0.01)
```

**Step 5 — annotate adducts and fragments with commonMZ rules.** CAMERA requires a plain `data.frame`, not a tibble, so wrap with `as.data.frame()`:

```{r step5}
#| eval: false
cam_pos <- findAdducts(cam_pos,
                       ppm        = 10,
                       mzabs      = 0.01,
                       multiplier = 3,
                       polarity   = "positive",
                       rules      = as.data.frame(rules_pos_clean))
```

`multiplier` controls the highest oligomer CAMERA will consider (e.g. `[2M+H]+`, `[3M+H]+`). `ppm` and `mzabs` set the matching tolerance; CAMERA applies whichever window is wider, so for high-resolution data keep `mzabs` small (0.005–0.01) and rely on `ppm`.

------------------------------------------------------------------------

The code below runs this pipeline on `mm14`, the xcmsSet bundled with CAMERA (134 peaks, single-sample, positive mode), to produce the outputs shown here.

```{r run-example}
#| code-fold: true
library(CAMERA)

data("mm14")                                          # xcmsSet from CAMERA package

xsa        <- xsAnnotate(mm14, polarity = "positive")
xsaF       <- groupFWHM(xsa, perfwhm = 0.6, intval = "into")
xsaI       <- findIsotopes(xsaF, ppm = 10, mzabs = 0.01, intval = "into")
cam_result <- findAdducts(xsaI,
                          ppm        = 10,
                          mzabs      = 0.01,
                          multiplier = 3,
                          polarity   = "positive",
                          rules      = as.data.frame(rules_pos_clean))
```

## Annotated peak table

`getPeaklist()` returns one row per feature with annotation columns appended. The `adduct` column shows the best-matching rule for each feature; features with no match are left blank. Features with the same pcgroup number are assumed to originate from the same parent moelcule.

```{r peaklist-table}
#| code-fold: true
#| eval: false
peaklist <- getPeaklist(cam_result)
peaklist %>%
  select(mz, rt, any_of(c("isotopes", "adduct", "pcgroup")), starts_with("sample")) %>%
  mutate(pcgroup = as.integer(pcgroup)) %>%
  arrange(pcgroup, mz) %>%
  mutate(mz = round(mz, 4), rt = round(rt, 1))
```

```{r peaklist-table-dt}
#| echo: false
peaklist <- getPeaklist(cam_result)
peaklist %>%
  select(mz, rt, any_of(c("isotopes", "adduct", "pcgroup")), starts_with("sample")) %>%
  mutate(pcgroup = as.integer(pcgroup)) %>%
  arrange(pcgroup, mz) %>%
  mutate(mz = round(mz, 4), rt = round(rt, 1)) %>%
  DT::datatable(rownames = FALSE, options = list(pageLength = 15, scrollX = TRUE))
```

## Which rules fired?

After annotation, count how often each rule was assigned. Rules that fired zero times are either absent from this matrix or could not be resolved within the tolerance.

We could do a pie chart:

```{r rules-freq}
#| code-fold: true
camera_pie(cam_result)
```

Or a fancy Sankey chart:

```{r rules-sankey}
#| code-fold: true
camera_sankey(cam_result)
```
