---
title: "Reading isotopic fine structure"
subtitle: "Why the M+2 peak is not one peak"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    fig-format: svg
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Reading isotopic fine structure}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(dplyr)
library(purrr)
library(ggplot2)

BLUE <- "#2a78d6"
COL  <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
         S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
```

## Chlorine's M+2: an abundance effect

Chlorine has two stable isotopes: ³⁵Cl (75.8% natural abundance, monoisotopic) and
³⁷Cl (24.2%). A molecule with two chlorines -- like **diclofenac** (C₁₄H₁₁Cl₂NO₂) --
has roughly a 37% chance of carrying exactly one ³⁷Cl, placing that molecule at M+2.
The monoisotopic M peak is simultaneously thinned out as ions scatter across M+1,
M+2, M+3 and M+4 -- so M+2 ends up at ~65% of M, an excess so large that no
fine-structure argument is required to recognize it.

```{r diclofenac-coarse}
#| fig-cap: "Diclofenac's isotope pattern at unit resolution. Two chlorines are unmissable without any fine-structure argument at all."
pat_dic <- isotope_fine_pattern("C14H11Cl2NO2")
base_dic <- pat_dic %>% filter(label == "") %>% pull(mz)

coarse_dic <- pat_dic %>%
  mutate(nominal = round(mz - base_dic)) %>%
  group_by(nominal) %>%
  summarise(mz = base_dic + first(nominal), abundance = sum(abundance), .groups = "drop")

ggplot(coarse_dic, aes(mz, abundance)) +
  geom_segment(aes(xend = mz, y = 0, yend = abundance), colour = BLUE, linewidth = 2.2,
               lineend = "round") +
  geom_text(data = filter(coarse_dic, abundance > 1), aes(label = paste0("M+", nominal)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z (Da)", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

That coarse view is `isotope_fine_pattern()`'s output summed by nominal mass --
the same number every isotope-pattern calculator reports. M+2 at ~65% follows
directly from ³⁷Cl's 24.2% natural abundance.

## Why different elements land at different exact masses

A "nominal +2" peak contains any isotopologue whose mass rounds to 2 Da above the
monoisotopic peak. But exact masses differ: nuclear binding energy -- the energy
released when protons and neutrons form a nucleus -- varies between nuclei, so the
heavy isotope of each element sits at its own characteristic offset from the nearest
integer.

| Isotopologue | Shift from monoisotopic (Da) | Offset from nominal +2 (mDa) |
|---|---:|---:|
| ¹³C + ¹³C | +2.0067 | +6.7 |
| ¹⁸O | +2.0042 | +4.2 |
| ³⁷Cl | +1.9971 | −2.9 |
| ³⁴S | +1.9958 | −4.2 |

The 10.9 mDa gap between ¹³C¹³C and ³⁴S is larger than the mass accuracy of any
modern high-resolution instrument: they are at genuinely different masses. Whether
your spectrum actually resolves them as two peaks rather than one depends on
resolving power -- which the sections below work through for a concrete example.

## When the coarse pattern hides the answer

**Methionine**, C₅H₁₁NO₂S, is about as ordinary as a metabolite gets, one of the
twenty proteinogenic amino acids, present in every untargeted metabolomics urine or
plasma run. It also carries one sulfur, and sulfur's own +2 isotope, ³⁴S, sits at
4.25% abundance, two orders of magnitude more abundant than the +1 isotopes of the
CHNO elements. Nothing about a coarse M+1/M+2 barcode makes that obvious the way
diclofenac's two chlorines are obvious.

```{r methionine-coarse}
#| fig-cap: "Methionine's coarse isotope pattern. M+2 is bigger than five carbons alone would ever produce - but the bars alone don't say why."
pat_met <- isotope_fine_pattern("C5H11NO2S")
base_met <- pat_met %>% filter(label == "") %>% pull(mz)

coarse_met <- pat_met %>%
  mutate(nominal = round(mz - base_met)) %>%
  group_by(nominal) %>%
  summarise(mz = base_met + first(nominal), abundance = sum(abundance), .groups = "drop")

ggplot(coarse_met, aes(mz, abundance)) +
  geom_segment(aes(xend = mz, y = 0, yend = abundance), colour = BLUE, linewidth = 2.2,
               lineend = "round") +
  geom_text(data = filter(coarse_met, abundance > 1), aes(label = paste0("M+", nominal)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z (Da)", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Five carbons predict an M+1 of roughly `5 * 1.07% ≈ 5.4%` of M, reasonably close to
what the chart shows, since carbon dominates M+1 for any CHNOS compound this size.
But M+2 at `r round(coarse_met %>% filter(nominal==2) %>% pull(abundance), 1)`% is the number
that should make you stop: naively, M+2 "ought" to come from picking any two of
those five carbons to both be ¹³C, which combinatorics puts at only
`choose(5,2) * 0.0107^2 * 100 ≈ 0.11%`, **roughly 40 times smaller** than what is
actually observed. Something other than double-¹³C is producing most of that peak,
and the coarse pattern alone cannot tell you what. That gap between the naive
¹³C-only prediction and the observed M+2 height is itself a diagnostic: whenever
M+2 badly overshoots what carbon alone predicts, suspect S, Cl, Br, Si, or a metal
before anything else; that is exactly the reasoning
[`contaminants_+.tsv`](contaminants_+.tsv)'s metal-complex entries were built on.

## What is actually inside methionine's M+2

`isotope_fine_pattern()` reports each isotopologue separately, at its own exact mass,
labelled by which isotopes produced it:

```{r met-fine-table}
pat_met %>% arrange(desc(abundance)) %>%
  mutate(mz = round(mz, 5), abundance = round(abundance, 4),
         label = ifelse(label == "", "M (monoisotopic)", label)) %>%
  knitr::kable(caption = "Every isotopologue of methionine above the default 0.01% threshold.")
```

Plotted on the m/z axis (the same Da scale your spectrum viewer uses, no milli-Da
rescaling), the M+1 and M+2 regions each split into several distinct peaks:

```{r met-m1}
#| fig-cap: "Methionine, M+1: five candidate isotope substitutions, spread across 0.009 Da. Carbon dominates, as it does for almost every organic ion."
plot_level <- function(pattern, level, title) {
  base <- pattern %>% filter(label == "") %>% pull(mz)
  d <- pattern %>% mutate(nominal = round(mz - base)) %>% filter(nominal == level) %>%
    mutate(element = sub(",.*", "", label), element = sub("^[0-9]+", "", element),
           col = ifelse(label == "" | grepl(",", label), "combo", element),
           col = ifelse(col %in% names(COL), col, "combo"))
  ggplot(d, aes(mz, abundance, colour = col)) +
    geom_segment(aes(xend = mz, y = 0, yend = abundance), linewidth = 2.2, lineend = "round") +
    geom_text(data = slice_max(d, abundance, n = min(6, nrow(d))),
              aes(label = label), angle = 90, hjust = -0.15, size = 3.1, colour = "grey20") +
    scale_colour_manual(values = COL, guide = "none") +
    scale_x_continuous(labels = scales::label_number(accuracy = 0.0001)) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .35))) +
    labs(x = "m/z (Da)", y = "% of M (monoisotopic)", title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
plot_level(pat_met, 1, "Methionine M+1")
```

```{r met-m2}
#| fig-cap: "Methionine, M+2: 34S alone outweighs every carbon-only combination roughly 40-fold, exactly matching the excess spotted in the coarse pattern above."
plot_level(pat_met, 2, "Methionine M+2")
```

Two things the coarse bar chart could never show are visible here. First,
**sulfur's own +2 isotope physically sits apart from every carbon-based
explanation**: `13C, 13C` (two different carbons, both ¹³C) sits
`r { d2 <- pat_met %>% mutate(nominal = round(mz - base_met)) %>% filter(nominal == 2); round(abs((d2 %>% filter(label=="13C, 13C") %>% pull(mz)) - (d2 %>% filter(label=="34S") %>% pull(mz))), 5) }`
Da away from `34S`, a large enough gap that they are two genuinely different
masses, not two names for the same peak. Second, the height difference *is* the
sulfur signal: `34S` towers over every combination of carbons, nitrogens and
oxygens that could also reach +2, which is the fine-structure version of the
40-fold excess already spotted in the coarse pattern, now with a specific
competing hypothesis (¹³C¹³C) at a specific, different mass instead of a
vague "something else."

## Whether you'd actually see it depends on resolving power

A mass difference on paper is not the same as two resolved peaks on a
chromatogram. Whether `34S` and `13C, 13C` show up as one blob or two depends
entirely on the resolving power available at that m/z. Resolving power is not
a fixed number: Orbitraps are quoted at a reference m/z (typically 200) and lose
resolving power roughly as `1/sqrt(m/z)` away from it. `isotope_profile()` turns
that into a picture rather than an abstraction, by summing a Gaussian peak of the
right width for each isotopologue and showing what the region would actually look
like at a chosen resolving power:

```{r sim-profile}
#| fig-cap: "Methionine's M+2 region simulated at three Orbitrap resolving powers (quoted @ m/z 200). Vertical ticks mark where each candidate actually sits. At 15,000 the shoulder is invisible; by 60,000 it is unambiguous."
R_at_mz <- function(mz, R_ref, mz_ref = 200) R_ref * sqrt(mz_ref / mz)

Rs <- c(15000, 60000, 120000)
R_label <- function(r) sprintf("R = %s @200", format(r, big.mark = ",", trim = TRUE))
pat_met2 <- pat_met %>% mutate(nominal = round(mz - base_met)) %>% filter(nominal == 2)

prof <- map_dfr(Rs, function(r)
  isotope_profile(pat_met2, resolution = R_at_mz(mean(pat_met2$mz), r)) %>%
    mutate(R = R_label(r)))
ticks <- pat_met2 %>% filter(abundance > 0.02)

ggplot(prof, aes(mz, intensity)) +
  geom_area(fill = BLUE, alpha = .25) +
  geom_line(colour = BLUE, linewidth = .8) +
  geom_rug(data = ticks, aes(x = mz), inherit.aes = FALSE, sides = "b", colour = "grey30", linewidth = .6) +
  facet_wrap(~ factor(R, levels = map_chr(Rs, R_label)), ncol = 1, scales = "free_y") +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.0001)) +
  labs(x = "m/z (Da)", y = "simulated intensity (% of M)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(hjust = 0))
```

At 15,000 the ³⁴S peak is a single, slightly lopsided blob; the ¹³C¹³C shoulder is
in there, but you cannot see it. By 60,000 it is an unmistakable second peak. This
is the actual lesson of fine structure: it is not a yes/no property of a molecule,
it is a question you can only answer once you know both *what* could be there and
*whether your instrument could show it to you*.

## Where to go from here

This document covered the *why*: the arithmetic behind a mass defect and why
resolving power changes what you can conclude from a spectrum. Two
companion vignettes turn that understanding into a workflow you can run on your own
data:

* [**Simulating isotope fine structure for a formula**](isotope-fine-structure-tool.html):
  given a candidate formula, get its full fine-structure table, a resolving-power
  calculator for every pair of candidates, and a simulated profile at your own
  instrument's settings.
* [**Looking up an unexplained mass difference**](mass-difference-lookup.html):
  the opposite problem: you have a delta between two peaks and no formula yet, and
  want to search it against every known adduct, in-source fragment and
  repeating-unit difference in commonMZ at once.
