---
title: "Simulating isotope fine structure for a formula"
subtitle: "Problem: I have a candidate formula. What does its fine structure look like, and could I resolve it?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    fig-format: svg
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Simulating isotope fine structure for a formula}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(commonMZ)
library(dplyr)
library(purrr)
library(ggplot2)
library(plotly)

BLUE <- "#2a78d6"
COL  <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
         S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
```

This is the practical companion to
[*Reading isotopic fine structure*](isotope-fine-structure-explained.html), which
explains the concept using methionine as a worked example. Here the goal is a
recipe you can run on any formula of your own: build the fine structure, check
whether your instrument would resolve the candidates that matter, and see the
simulated peak shape at your own resolving power.

## Step 1 --- build the fine structure

Swap in your own candidate formula here:

```{r pattern}
formula <- "C14H11Cl2NO2"   # diclofenac -- replace with your own candidate

pat <- isotope_fine_pattern(formula)
pat %>% arrange(mz) %>%
  mutate(mz = round(mz, 5), abundance = round(abundance, 4),
         label = ifelse(label == "", "M (monoisotopic)", label)) %>%
  knitr::kable(caption = paste("Every isotopologue of", formula, "above the default 0.01% threshold."))
```

Grouping by nominal mass reproduces the coarse M/M+1/M+2 pattern any calculator
would give you --- diclofenac's two chlorines make this one unambiguous even at unit
resolution:

```{r coarse}
#| fig-cap: "Diclofenac's coarse isotope pattern: two chlorines produce a textbook M/M+2/M+4 barcode."
base_mz <- pat %>% filter(label == "") %>% pull(mz)
coarse <- pat %>% mutate(nominal = round(mz - base_mz)) %>%
  group_by(nominal) %>%
  summarise(mz = base_mz + first(nominal), abundance = sum(abundance), .groups = "drop")

ggplot(coarse, aes(mz, abundance)) +
  geom_segment(aes(xend = mz, y = 0, yend = abundance), colour = BLUE, linewidth = 2.2,
               lineend = "round") +
  geom_text(data = filter(coarse, abundance > 1), aes(label = paste0("M+", nominal)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z (Da)", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Step 2 --- zoom into one nominal level

Plotting the isotopologues of a single nominal level directly on the m/z axis
shows what is really inside it --- here M+2, where a 37Cl isotopologue dominates so
completely that a small 13C,13C shoulder is almost invisible on a linear scale:

```{r zoom}
#| fig-cap: "Diclofenac M+2: two chlorines dominate so completely that the tiny 13C,13C shoulder barely registers, even though it sits at a perfectly resolvable distance."
plot_level <- function(pattern, level, title) {
  base <- pattern %>% filter(label == "") %>% pull(mz)
  d <- pattern %>% mutate(nominal = round(mz - base)) %>% filter(nominal == level) %>%
    mutate(element = sub(",.*", "", label), element = sub("^[0-9]+", "", element),
           col = ifelse(label == "" | grepl(",", label), "combo", element),
           col = ifelse(col %in% names(COL), col, "combo"))
  ggplot(d, aes(mz, abundance, colour = col)) +
    geom_segment(aes(xend = mz, y = 0, yend = abundance), linewidth = 2.2, lineend = "round") +
    geom_text(data = slice_max(d, abundance, n = min(6, nrow(d))),
              aes(label = label), angle = 90, hjust = -0.15, size = 3.1, colour = "grey20") +
    scale_colour_manual(values = COL, guide = "none") +
    scale_x_continuous(labels = scales::label_number(accuracy = 0.0001)) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .35))) +
    labs(x = "m/z (Da)", y = "% of M (monoisotopic)", title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
plot_level(pat, 2, paste(formula, "M+2"))
```

## Step 3 --- would your instrument resolve the candidates?

For every pair of isotopologues at a nominal level the table below reports the
resolving power (Orbitrap convention, quoted at m/z 200) at which the two Gaussian
peaks would show a ~50% valley between them --- enough for a centroid algorithm to
find two separate peaks. Two Gaussian peaks of equal height separated by 1.5 × FWHM
produce a valley at about 50% of peak height; translating that to the R@200 Orbitrap
convention scales as `1/sqrt(m/z)`:

```{r resolving-power}
R_at_mz <- function(mz, R_ref, mz_ref = 200) R_ref * sqrt(mz_ref / mz)

separations <- function(pattern, level, mz_ref = 200, min_abundance = 0.01) {
  base <- pattern %>% filter(label == "") %>% pull(mz)
  d <- pattern %>% mutate(nominal = round(mz - base)) %>%
    filter(nominal == level, abundance >= min_abundance) %>%
    mutate(label = ifelse(label == "", "M", label))
  if (nrow(d) < 2) return(tibble())
  ix <- combn(seq_len(nrow(d)), 2)
  map_dfr(seq_len(ncol(ix)), function(i) {
    ra <- d[ix[1, i], ]; rb <- d[ix[2, i], ]
    sep_da  <- abs(ra$mz - rb$mz)
    m_avg   <- (ra$mz + rb$mz) / 2
    Rreq    <- 1.5 * m_avg / sep_da              # R at this m/z for ~50% valley
    Rref_req <- Rreq / sqrt(mz_ref / m_avg)      # translate to @mz_ref convention
    tibble(a = ra$label, b = rb$label, abundance_a_pct = round(ra$abundance, 3),
           abundance_b_pct = round(rb$abundance, 3), separation_Da = round(sep_da, 5),
           R_at_mz = round(Rreq), R_at200 = round(Rref_req))
  }) %>% arrange(R_at200)
}

sep2 <- separations(pat, 2)
sep2 %>%
  knitr::kable(col.names = c("a", "b", "abundance a %", "abundance b %",
                              "separation (Da)", "R at this m/z", "R @200"),
               caption = "Every pair of M+2 candidates above 0.01% abundance. R@200 is the Orbitrap resolving power (quoted at m/z 200) at which the two peaks show a ~50% valley --- enough for centroiding to find two peaks.")
```

```{r sep-text, results = "asis"}
top <- sep2 %>% filter(a == "37Cl" | b == "37Cl") %>% slice_min(R_at200, n = 1)
cat(sprintf("`37Cl` needs only R ~ %s @200 to show a distinct peak from its closest competitor (`%s`) --- comfortably within reach on any current Orbitrap.",
            format(top$R_at200, big.mark = ","), ifelse(top$a == "37Cl", top$b, top$a)))
```

## Step 4 --- simulate the profile at your own resolving power

Numbers on a table are one thing; seeing whether two peaks actually merge is more
convincing. `isotope_profile()` sums a Gaussian peak of the right width for each
candidate --- pass it the *actual* resolving power at this m/z, converting from an
Orbitrap's `R_ref @ 200` setting with `R_at_mz()` first:

```{r sim-profile}
Rs <- c(30000, 60000, 120000)
R_label <- function(r) sprintf("R = %s @200", format(r, big.mark = ",", trim = TRUE))
pat2 <- pat %>% mutate(nominal = round(mz - base_mz)) %>% filter(nominal == 2)

prof <- map_dfr(Rs, function(r)
  isotope_profile(pat2, resolution = R_at_mz(mean(pat2$mz), r)) %>%
    mutate(R = R_label(r), R_order = r))
ticks <- pat2 %>% filter(abundance > 0.5)

R_levels <- map_chr(Rs, R_label)
panels <- map(seq_along(Rs), function(i) {
  d <- prof %>% filter(R == R_levels[i])
  plot_ly(d, x = ~mz, y = ~intensity, type = "scatter", mode = "lines",
          fill = "tozeroy", fillcolor = paste0(BLUE, "40"),
          line = list(color = BLUE, width = 1.2),
          showlegend = FALSE,
          hovertemplate = "%{x:.4f} Da<br>%{y:.3f}%<extra></extra>") %>%
    add_trace(data = ticks, x = ~mz, y = 0, type = "scatter", mode = "markers",
              marker = list(symbol = "line-ns-open", size = 10, color = "grey30"),
              showlegend = FALSE, hoverinfo = "none") %>%
    layout(annotations = list(
      list(x = 0.01, y = 0.97, text = R_levels[i], xref = "paper", yref = "paper",
           xanchor = "left", yanchor = "top", showarrow = FALSE,
           font = list(size = 12))
    ))
})

subplot(panels, nrows = length(Rs), shareX = TRUE, titleY = FALSE) %>%
  layout(
    xaxis  = list(title = "m/z (Da)", tickformat = ".4f"),
    yaxis  = list(title = "simulated intensity (% of M)"),
    yaxis2 = list(title = ""),
    yaxis3 = list(title = ""),
    margin = list(l = 60)
  )
```

Even at the lowest of these three resolving powers the two chlorines are already
visually obvious as a separate hump --- consistent with the table above, and the
kind of case where fine structure confirms what the coarse pattern already showed
rather than revealing something it hid. For a case where fine structure genuinely
changes the conclusion, see methionine's hidden sulfur in
[*Reading isotopic fine structure*](isotope-fine-structure-explained.html).

## Checklist

1. **Build the pattern** with `isotope_fine_pattern(formula)` and check the coarse
   sum (`group_by` nominal mass) against what carbon count alone predicts
   (`nC * 1.07%` for M+1; `choose(nC,2) * 0.0107^2 * 100%` for M+2). A large excess
   is the first signal that something heavier than carbon is present.
2. **Zoom into the nominal level** that shows the excess to see which specific
   isotopologues could produce it, and how far apart they sit in exact mass --- in
   Da, directly comparable to your spectrum's own axis.
3. **Check the gap against your resolving power** with `separations()` before
   concluding a peak is "clean" --- a shoulder that should be there but isn't
   resolved is not evidence it's absent, only evidence you didn't have the
   resolving power to see it.
4. **Simulate the profile** with `isotope_profile()` at your instrument's actual
   resolving power (converted from its `R @ reference m/z` spec with `R_at_mz()`)
   to see, rather than infer, whether the candidates would merge.

None of `plot_level()` or `separations()` are exported by commonMZ; only
`isotope_fine_pattern()` and `isotope_profile()` are. The two helpers above are
short enough to copy into your own analysis --- that's the point of showing their
full source here.
