---
title: "Looking up an isotopologue offset"
subtitle: "I see a satellite peak at a known distance from M+1 or M+2. Which element caused it?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Looking up an isotopologue offset}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(dplyr)
library(purrr)
library(DT)
library(htmltools)

## Elements to include in the reference table
REF_ELEMENTS <- c("C", "H", "N", "O", "S", "P", "F", "Cl", "Br", "I",
                   "Si", "B", "Se", "Fe", "Ca", "K", "Na", "Mg", "Zn", "Cu")
```

When you measure a high-resolution spectrum and see a small satellite peak a few
millidaltons away from the main M+1 or M+2 peak, the question is: what isotope
substitution sits there? This page provides a precomputed reference of every
isotopologue offset for common elements and a search that works exactly like the
[mass-difference lookup](mass-difference-lookup.html): enter an observed offset and
a tolerance, click **Search**.

Offsets are reported in two ways:

* **offset from M** (Da): the exact mass shift from the monoisotopic peak, i.e.
  the number that `isotope_fine_pattern()` gives you in its `mz` column minus the
  monoisotopic `mz`.
* **offset from ¹³C** (for M+1) or **offset from ¹³C,¹³C** (for M+2): how far
  the isotopologue sits from the dominant carbon reference within that nominal level.
  This is what you read off the m/z axis when you zoom into the M+1 or M+2 region
  of a carbon-containing compound.

```{r build-ref}
## Build one row per element × isotopologue combination for M+1 and M+2
ref_raw <- map_dfr(REF_ELEMENTS, function(el) {
  p <- tryCatch(isotope_fine_pattern(el), error = function(e) NULL)
  if (is.null(p) || nrow(p) == 0) return(NULL)
  base <- p %>% filter(label == "") %>% pull(mz)
  if (length(base) == 0) return(NULL)
  p %>%
    filter(label != "") %>%
    mutate(
      element  = el,
      nominal  = round(mz - base),
      offset_M = round(mz - base, 6)
    ) %>%
    filter(nominal %in% 1:2) %>%
    select(element, level = nominal, isotopologue = label, offset_M, abundance)
})

## Reference masses: 13C for M+1, "13C, 13C" for M+2
c13_m1 <- ref_raw %>%
  filter(element == "C", level == 1, isotopologue == "13C") %>%
  pull(offset_M)
c13c13_m2 <- ref_raw %>%
  filter(element == "C", level == 2, isotopologue == "13C, 13C") %>%
  pull(offset_M)

offsets <- ref_raw %>%
  mutate(
    offset_from_ref = case_when(
      level == 1 & length(c13_m1)    > 0 ~ round(offset_M - c13_m1[1],    6),
      level == 2 & length(c13c13_m2) > 0 ~ round(offset_M - c13c13_m2[1], 6),
      TRUE ~ NA_real_
    ),
    reference = if_else(level == 1, "13C", "13C,13C"),
    level     = paste0("M+", level),
    abundance = round(abundance, 4)
  ) %>%
  arrange(level, offset_from_ref) %>%
  select(level, element, isotopologue, `offset from M (Da)` = offset_M,
         reference, `offset from ref (Da)` = offset_from_ref,
         `abundance (% of M)` = abundance)
```

## Search by observed offset

Type the offset you measured in the spectrum and a tolerance in ppm (of the offset
value, the same convention as `mz_diff_lookup()`). Click **Search**, then **Clear**
to reset.

```{r offset-interactive}
## DataTables core version bundled with DT, for the CDN fallback below
probe_widget      <- datatable(data.frame(x = 1))
DT_CORE_VERSION   <- Filter(function(d) d$name == "dt-core",
                             probe_widget$dependencies)[[1]]$version

tagList(
  tags$div(
    style = "display:flex; gap:1em; align-items:end; margin-bottom:0.75em; flex-wrap:wrap;",
    tags$div(
      tags$label("offset (Da)", `for` = "iso_off"),
      tags$br(),
      tags$input(id = "iso_off", type = "number", step = "any",
                 class = "form-control", style = "width:12em;")
    ),
    tags$div(
      tags$label("tolerance (ppm)", `for` = "iso_ppm"),
      tags$br(),
      tags$input(id = "iso_ppm", type = "number", step = "any", value = "50",
                 class = "form-control", style = "width:10em;")
    ),
    tags$div(
      tags$label("column to search", `for` = "iso_col"),
      tags$br(),
      tags$select(id = "iso_col", class = "form-control", style = "width:16em;",
        tags$option(value = "3", "offset from M (Da)",     selected = NA),
        tags$option(value = "5", "offset from ref (Da)")
      )
    ),
    tags$button("Search", id = "iso_go",    type = "button", class = "btn btn-primary"),
    tags$button("Clear",  id = "iso_clear", type = "button", class = "btn btn-outline-secondary")
  ),
  datatable(offsets, rownames = FALSE, elementId = "iso_table",
            options = list(pageLength = 15, order = list(list(5, "asc")))),
  tags$script(HTML(sprintf("
    if (typeof $ === 'undefined' || typeof $.fn.dataTable === 'undefined') {
      document.write('<script src=\"https://cdn.datatables.net/%s/js/jquery.dataTables.min.js\"><\\/script>');
    }
    $(document).ready(function() {
      var $tbl = $('#iso_table').find('table').first();
      if (!$tbl.length) return;
      var dt = $tbl.DataTable();
      var target = null, tolDa = null, colIdx = 3;

      $.fn.dataTable.ext.search.push(function(settings, data, dataIndex) {
        if (settings.nTable !== $tbl.get(0)) return true;
        if (target === null) return true;
        var val = parseFloat(data[colIdx]);
        return !isNaN(val) && Math.abs(val - target) <= tolDa;
      });

      $('#iso_go').on('click', function() {
        colIdx = parseInt($('#iso_col').val()) - 1;
        var off = parseFloat(document.getElementById('iso_off').value);
        var ppm = parseFloat(document.getElementById('iso_ppm').value);
        if (isNaN(off) || isNaN(ppm)) { target = null; }
        else { target = off; tolDa = Math.abs(off) * ppm / 1e6; }
        dt.draw();
      });
      $('#iso_clear').on('click', function() {
        document.getElementById('iso_off').value = '';
        target = null;
        dt.draw();
      });
    });
  ", DT_CORE_VERSION)))
)
```

The search re-filters only when you click **Search**. Choose whether to search by
**offset from M** (what `isotope_fine_pattern()` gives directly) or **offset from
ref** (what you read off the m/z axis when zoomed into the M+1 or M+2 region).

## How to use these offsets

1. Zoom into the M+1 or M+2 region of your spectrum and measure the distance
   between the main carbon peak (¹³C at M+1; ¹³C,¹³C at M+2) and the satellite.
2. Enter that distance in the **offset from ref** search above with a tolerance
   matching your instrument's mass accuracy.
3. Check abundance: if the satellite is smaller than the predicted abundance, the
   element may be present but not resolved; if it is absent entirely, the element
   is not present (or the resolving power was insufficient).

For a formula-specific analysis — how large the satellite should be, what resolving
power you need to see it — see [*Isotope fine structure*](isotope-fine-structure.html).
