---
title: "Reading isotopic fine structure"
subtitle: "Why the M+2 peak is not one peak"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    fig-format: svg
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svg
vignette: >
  %\VignetteIndexEntry{Reading isotopic fine structure}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(commonMZ)
library(dplyr)
library(ggplot2)

BLUE <- "#2a78d6"
COL  <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
         S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
```

## The question this document answers

Every ion table in commonMZ — the adducts, the contaminants, the repeating units —
describes a *monoisotopic* mass. But nothing you actually measure is monoisotopic
only: every real peak drags a family of satellites behind it, one nominal mass unit
up, then two, then three. Most of the time you glance at the M+1 and M+2 satellites,
eyeball their height relative to M, and move on — "small M+1, must be a couple of
carbons; big M+2, must be sulfur or chlorine."

That shortcut works because of a coincidence that is worth making explicit: **at
unit mass resolution, M+2 looks like a single peak, but it is actually the sum of
several physically distinct isotopologues that happen to round to the same nominal
mass.** ¹³C¹³C, ³⁴S, ¹⁸O, ³⁷Cl and ⁸¹Br can all land one M+2 satellite depending on
what the molecule contains, and they do not have the same exact mass — each is off
from the true integer by its own small mass defect, typically a few thousandths of
a Da. At the resolving power routinely available on an Orbitrap, several of those
gaps are wide enough to *see*: the "M+2 peak" resolves into two or three peaks that
individually confirm which isotopes are actually present.

This document builds that fine structure with commonMZ's own
`isotope_fine_pattern()`, which wraps [enviPat](https://cran.r-project.org/package=enviPat)
(Loos, Singer & Meringer, *Metabolites* 2015) — the same engine used in
[xcmsVisGUI](https://github.com/stanstrup/xcmsVisGUI) — so every mass and natural
abundance below is enviPat's, never typed by hand. All masses and differences are
reported in Da, at full precision, so you can compare them directly against the m/z
axis of your own spectrum rather than mentally rescaling from milli-Da.

Once you have the concept, two companion vignettes turn it into a workflow:
[*Simulating isotope fine structure for a formula*](isotope-fine-structure-tool.html)
for "I have a candidate formula, what does its fine structure look like?", and
[*Looking up an unexplained mass difference*](mass-difference-lookup.html) for the
opposite situation, where you have a delta between two peaks and no formula yet.

## The coarse pattern everyone learns first

Start with a compound whose isotope pattern needs no explaining: **diclofenac**,
C₁₄H₁₁Cl₂NO₂, an anti-inflammatory drug common enough as a urinary xenobiotic that
it turns up in untargeted metabolomics data. Two chlorines give it a textbook
barcode.

```{r diclofenac-coarse}
#| fig-cap: "Diclofenac's isotope pattern at unit resolution. Two chlorines are unmissable without any fine-structure argument at all."
pat_dic <- isotope_fine_pattern("C14H11Cl2NO2")
base_dic <- pat_dic$mz[pat_dic$label == ""]

coarse_dic <- pat_dic %>%
  mutate(nominal = round(mz - base_dic)) %>%
  group_by(nominal) %>%
  summarise(mz = base_dic + first(nominal), abundance = sum(abundance), .groups = "drop")

ggplot(coarse_dic, aes(mz, abundance)) +
  geom_segment(aes(xend = mz, y = 0, yend = abundance), colour = BLUE, linewidth = 2.2,
               lineend = "round") +
  geom_text(data = filter(coarse_dic, abundance > 1), aes(label = paste0("M+", nominal)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z (Da)", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

That coarse view is just `isotope_fine_pattern()`'s output summed by nominal mass —
the same number every isotope-pattern calculator reports. That is genuinely all
you need here — M+2 at 65% of M is such an overwhelming excess over what carbon
alone could produce that no fine-structure argument is required. Two chlorines
announce themselves in a glance.

The interesting cases are the ones where the coarse pattern is *not* dramatic
enough to read at a glance, and the only way to be sure what is in a peak is to
look inside it.

## When the coarse pattern hides the answer

**Methionine**, C₅H₁₁NO₂S, is about as ordinary as a metabolite gets — one of the
twenty proteinogenic amino acids, present in every untargeted metabolomics urine or
plasma run. It also carries one sulfur, and sulfur's own +2 isotope, ³⁴S, sits at
4.25% abundance — two orders of magnitude more abundant than the +1 isotopes of the
CHNO elements. Nothing about a coarse M+1/M+2 barcode makes that obvious the way
diclofenac's two chlorines are obvious.

```{r methionine-coarse}
#| fig-cap: "Methionine's coarse isotope pattern. M+2 is bigger than five carbons alone would ever produce - but the bars alone don't say why."
pat_met <- isotope_fine_pattern("C5H11NO2S")
base_met <- pat_met$mz[pat_met$label == ""]

coarse_met <- pat_met %>%
  mutate(nominal = round(mz - base_met)) %>%
  group_by(nominal) %>%
  summarise(mz = base_met + first(nominal), abundance = sum(abundance), .groups = "drop")

ggplot(coarse_met, aes(mz, abundance)) +
  geom_segment(aes(xend = mz, y = 0, yend = abundance), colour = BLUE, linewidth = 2.2,
               lineend = "round") +
  geom_text(data = filter(coarse_met, abundance > 1), aes(label = paste0("M+", nominal)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z (Da)", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Five carbons predict an M+1 of roughly `5 * 1.07% ≈ 5.4%` of M — reasonably close to
what the chart shows, since carbon dominates M+1 for any CHNOS compound this size.
But M+2 at `r round(coarse_met$abundance[coarse_met$nominal==2],1)`% is the number
that should make you stop: naively, M+2 "ought" to come from picking any two of
those five carbons to both be ¹³C, which combinatorics puts at only
`choose(5,2) * 0.0107^2 * 100 ≈ 0.11%` — **roughly 40 times smaller** than what is
actually observed. Something other than double-¹³C is producing most of that peak,
and the coarse pattern alone cannot tell you what. That gap between the naive
¹³C-only prediction and the observed M+2 height is itself a diagnostic: whenever
M+2 badly overshoots what carbon alone predicts, suspect S, Cl, Br, Si, or a metal
before anything else — which is exactly the reasoning
[`contaminants_+.tsv`](contaminants_+.tsv)'s metal-complex entries were built on.

## What is actually inside methionine's M+2

`isotope_fine_pattern()` doesn't stop at the nominal sum — every row is one
isotopologue, at its own exact mass, labelled by which isotopes produced it:

```{r met-fine-table}
pat_met %>% arrange(desc(abundance)) %>%
  mutate(mz = round(mz, 5), abundance = round(abundance, 4),
         label = ifelse(label == "", "M (monoisotopic)", label)) %>%
  knitr::kable(caption = "Every isotopologue of methionine above the default 0.01% threshold, from enviPat.")
```

Plotted directly on the m/z axis — the same Da scale your spectrum viewer uses, no
milli-Da rescaling — the M+1 and M+2 regions each split into several distinct
peaks:

```{r met-m1}
#| fig-cap: "Methionine, M+1: five candidate isotope substitutions, spread across 0.009 Da. Carbon dominates, as it does for almost every organic ion."
plot_level <- function(pattern, level, title) {
  base <- pattern$mz[pattern$label == ""]
  d <- pattern %>% mutate(nominal = round(mz - base)) %>% filter(nominal == level) %>%
    mutate(element = sub(",.*", "", label), element = sub("^[0-9]+", "", element),
           col = ifelse(label == "" | grepl(",", label), "combo", element),
           col = ifelse(col %in% names(COL), col, "combo"))
  ggplot(d, aes(mz, abundance, colour = col)) +
    geom_segment(aes(xend = mz, y = 0, yend = abundance), linewidth = 2.2, lineend = "round") +
    geom_text(data = slice_max(d, abundance, n = min(6, nrow(d))),
              aes(label = label), angle = 90, hjust = -0.15, size = 3.1, colour = "grey20") +
    scale_colour_manual(values = COL, guide = "none") +
    scale_x_continuous(labels = scales::label_number(accuracy = 0.0001)) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .35))) +
    labs(x = "m/z (Da)", y = "% of M (monoisotopic)", title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
plot_level(pat_met, 1, "Methionine M+1")
```

```{r met-m2}
#| fig-cap: "Methionine, M+2: 34S alone outweighs every carbon-only combination roughly 40-fold, exactly matching the excess spotted in the coarse pattern above."
plot_level(pat_met, 2, "Methionine M+2")
```

Two things the coarse bar chart could never show are visible here. First,
**sulfur's own +2 isotope physically sits apart from every carbon-based
explanation** — `13C, 13C` (two different carbons, both ¹³C) sits
`r { d2 <- pat_met %>% mutate(nominal = round(mz - base_met)) %>% filter(nominal == 2); round(abs(d2$mz[d2$label=="13C, 13C"] - d2$mz[d2$label=="34S"]), 5) }`
Da away from `34S`, a large enough gap that they are two genuinely different
masses, not two names for the same peak. Second, the height difference *is* the
sulfur signal: `34S` towers over every combination of carbons, nitrogens and
oxygens that could also reach +2, which is the fine-structure version of the
40-fold excess already spotted in the coarse pattern — now with a specific
competing hypothesis (¹³C¹³C) sitting at a specific, different mass, instead of a
vague "something else."

## Whether you'd actually see it depends on resolving power

A mass difference on paper is not the same as two resolved peaks on a
chromatogram. Whether `34S` and `13C, 13C` show up as one blob or two depends
entirely on the resolving power available at that m/z — and resolving power is not
a fixed number: Orbitraps are quoted at a reference m/z (typically 200) and lose
resolving power roughly as `1/sqrt(m/z)` away from it. `isotope_profile()` turns
that into a picture rather than an abstraction, by summing a Gaussian peak of the
right width for each isotopologue and showing what the region would actually look
like at a chosen resolving power:

```{r sim-profile}
#| fig-cap: "Methionine's M+2 region simulated at three Orbitrap resolving powers (quoted @ m/z 200). Vertical ticks mark where each candidate actually sits. At 15,000 the shoulder is invisible; by 60,000 it is unambiguous."
R_at_mz <- function(mz, R_ref, mz_ref = 200) R_ref * sqrt(mz_ref / mz)

Rs <- c(15000, 60000, 120000)
R_label <- function(r) sprintf("R = %s @200", format(r, big.mark = ",", trim = TRUE))
pat_met2 <- pat_met %>% mutate(nominal = round(mz - base_met)) %>% filter(nominal == 2)

prof <- bind_rows(lapply(Rs, function(r)
  isotope_profile(pat_met2, resolution = R_at_mz(mean(pat_met2$mz), r)) %>%
    mutate(R = R_label(r))))
ticks <- pat_met2 %>% filter(abundance > 0.02)

ggplot(prof, aes(mz, intensity)) +
  geom_area(fill = BLUE, alpha = .25) +
  geom_line(colour = BLUE, linewidth = .8) +
  geom_rug(data = ticks, aes(x = mz), inherit.aes = FALSE, sides = "b", colour = "grey30", linewidth = .6) +
  facet_wrap(~ factor(R, levels = vapply(Rs, R_label, "")), ncol = 1, scales = "free_y") +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.0001)) +
  labs(x = "m/z (Da)", y = "simulated intensity (% of M)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(hjust = 0))
```

At 15,000 the ³⁴S peak is a single, slightly lopsided blob — the ¹³C¹³C shoulder is
in there, but you cannot see it. By 60,000 it is an unmistakable second peak. This
is the actual lesson of fine structure: it is not a yes/no property of a molecule,
it is a question you can only answer once you know both *what* could be there and
*whether your instrument could show it to you*.

## Where to go from here

This document was about understanding *why* — the arithmetic behind a mass defect,
and why resolving power changes what you can conclude from a spectrum. Two
companion vignettes turn that understanding into a workflow you can run on your own
data:

* [**Simulating isotope fine structure for a formula**](isotope-fine-structure-tool.html) —
  given a candidate formula, get its full fine-structure table, a resolving-power
  calculator for every pair of candidates, and a simulated profile at your own
  instrument's settings.
* [**Looking up an unexplained mass difference**](mass-difference-lookup.html) —
  the opposite problem: you have a delta between two peaks and no formula yet, and
  want to search it against every known adduct, in-source fragment and
  repeating-unit difference in commonMZ at once.
