---
title: "Looking up an isotopologue offset"
subtitle: "I measure a gap between two fine-structure peaks. Which pair of elements explains it?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Looking up an isotopologue offset}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(dplyr)
library(purrr)
library(DT)
library(htmltools)

## Elements to include in the reference table
REF_ELEMENTS <- c("C", "H", "N", "O", "S", "P", "F", "Cl", "Br", "I",
                   "Si", "B", "Se", "Fe", "Ca", "K", "Na", "Mg", "Zn", "Cu")
```

When you zoom into the M+1 or M+2 region of a high-resolution spectrum you see
several peaks separated by a few millidaltons. The question is: which pair of
isotope substitutions produces the gap you just measured?

This page precomputes every pairwise separation between isotopologues of common
elements within the same nominal level. Enter the observed gap and a tolerance,
click **Search**, and the table returns every element pair that could explain it.

```{r build-ref}
## Build one row per element × isotopologue for M+1 and M+2
ref_raw <- map_dfr(REF_ELEMENTS, function(el) {
  p <- tryCatch(isotope_fine_pattern(el), error = function(e) NULL)
  if (is.null(p) || nrow(p) == 0) return(NULL)
  base <- p %>% filter(label == "") %>% pull(mz)
  if (length(base) == 0) return(NULL)
  p %>%
    filter(label != "") %>%
    mutate(
      element  = el,
      nominal  = round(mz - base),
      offset_M = round(mz - base, 6)
    ) %>%
    filter(nominal %in% 1:2) %>%
    select(element, level = nominal, isotopologue = label, offset_M, abundance)
})

## All pairwise separations within each nominal level
pairs <- map_dfr(split(ref_raw, ref_raw$level), function(d) {
  if (nrow(d) < 2) return(NULL)
  level_label <- paste0("M+", d$level[1])
  idx <- combn(nrow(d), 2)
  map_dfr(seq_len(ncol(idx)), function(i) {
    ra <- d[idx[1, i], ]; rb <- d[idx[2, i], ]
    tibble(
      level          = level_label,
      `isotopologue A` = paste0(ra$element, ": ", ra$isotopologue),
      `isotopologue B` = paste0(rb$element, ": ", rb$isotopologue),
      `difference (Da)` = round(abs(ra$offset_M - rb$offset_M), 6),
      `abundance A (% of M)` = round(ra$abundance, 4),
      `abundance B (% of M)` = round(rb$abundance, 4)
    )
  })
}) %>% arrange(level, `difference (Da)`)
```

## Search by observed separation

Enter the gap you measured between two fine-structure peaks and a tolerance. The
ppm is relative to the **reference m/z** (the parent ion's m/z, 300 Da by default),
not to the tiny millidalton gap itself — ppm of the gap would give an absurdly tight
window, whereas ppm of the parent ion is what your instrument's mass accuracy
actually refers to.

```{r offset-interactive}
#| echo: false
## DataTables core version bundled with DT, for the CDN fallback below
probe_widget    <- datatable(data.frame(x = 1))
DT_CORE_VERSION <- Filter(function(d) d$name == "dt-core",
                           probe_widget$dependencies)[[1]]$version

tagList(
  tags$div(
    style = "display:flex; gap:1em; align-items:end; margin-bottom:0.75em; flex-wrap:wrap;",
    tags$div(
      tags$label("observed gap (Da)", `for` = "iso_off"),
      tags$br(),
      tags$input(id = "iso_off", type = "number", step = "any",
                 class = "form-control", style = "width:12em;")
    ),
    tags$div(
      tags$label("tolerance (ppm)", `for` = "iso_ppm"),
      tags$br(),
      tags$input(id = "iso_ppm", type = "number", step = "any", value = "50",
                 class = "form-control", style = "width:10em;")
    ),
    tags$div(
      tags$label("reference m/z (Da)", `for` = "iso_refmz"),
      tags$br(),
      tags$input(id = "iso_refmz", type = "number", step = "any", value = "300",
                 class = "form-control", style = "width:10em;")
    ),
    tags$button("Search", id = "iso_go",    type = "button", class = "btn btn-primary"),
    tags$button("Clear",  id = "iso_clear", type = "button", class = "btn btn-outline-secondary")
  ),
  datatable(pairs, rownames = FALSE, elementId = "iso_table",
            options = list(pageLength = 15, order = list(list(3, "asc")))),
  tags$script(HTML(sprintf("
    if (typeof $ === 'undefined' || typeof $.fn.dataTable === 'undefined') {
      document.write('<script src=\"https://cdn.datatables.net/%s/js/jquery.dataTables.min.js\"><\\/script>');
    }
    $(document).ready(function() {
      var $tbl = $('#iso_table').find('table').first();
      if (!$tbl.length) return;
      var dt = $tbl.DataTable();
      var target = null, tolDa = null;

      $.fn.dataTable.ext.search.push(function(settings, data, dataIndex) {
        if (settings.nTable !== $tbl.get(0)) return true;
        if (target === null) return true;
        var val = parseFloat(data[3]);
        return !isNaN(val) && Math.abs(val - target) <= tolDa;
      });

      $('#iso_go').on('click', function() {
        var off = parseFloat(document.getElementById('iso_off').value);
        var ppm = parseFloat(document.getElementById('iso_ppm').value);
        var refEl = document.getElementById('iso_refmz');
        var ref   = refEl ? parseFloat(refEl.value) : NaN;
        if (isNaN(off) || isNaN(ppm)) { target = null; }
        else {
          target = off;
          tolDa  = (!isNaN(ref) && ref > 0 ? ref : Math.abs(off)) * ppm / 1e6;
        }
        dt.draw();
      });
      $('#iso_clear').on('click', function() {
        document.getElementById('iso_off').value = '';
        target = null;
        dt.draw();
      });
    });
  ", DT_CORE_VERSION)))
)
```

The table shows every pair of isotopologues within the same nominal level (M+1 or
M+2). Search by entering the gap you measured between two peaks in the fine
structure.

## How to use this table

1. Zoom into the M+1 or M+2 region of your spectrum and measure the separation
   between any two peaks you can resolve.
2. Enter that gap in the **observed gap** field with a tolerance matching your
   instrument's mass accuracy, and click **Search**.
3. Each returned row names the pair of isotopologues whose exact-mass positions are
   that far apart. Check the abundance columns: if one member of the pair is far too
   small to see at your current resolution, rule it out.

For a formula-specific analysis — how large each peak should be, and what resolving
power you need to separate them — see [*Isotope fine structure*](isotope-fine-structure.html).
