---
title: "Looking up an isotopologue offset"
subtitle: "I measure a gap between two fine-structure peaks. Which pair of elements explains it?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Looking up an isotopologue offset}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(dplyr)
library(purrr)
library(DT)
library(htmltools)

## Elements to include in the reference table
REF_ELEMENTS <- c("C", "H", "N", "O", "S", "P", "F", "Cl", "Br", "I",
                   "Si", "B", "Se", "Fe", "Ca", "K", "Na", "Mg", "Zn", "Cu")
```

When you zoom into the M+1 or M+2 region of a high-resolution spectrum you see
several peaks separated by a few millidaltons. The question is: which pair of
isotope substitutions produces the gap you just measured?

This page precomputes every pairwise separation between isotopologues of common
elements within the same nominal level. Enter the observed gap and a tolerance,
click **Search**, and the table returns every element pair that could explain it.

```{r build-ref}
## Build one row per element × isotopologue for M+1 and M+2
ref_raw <- map_dfr(REF_ELEMENTS, function(el) {
  p <- tryCatch(isotope_fine_pattern(el), error = function(e) NULL)
  if (is.null(p) || nrow(p) == 0) return(NULL)
  base <- p %>% filter(label == "") %>% pull(mz)
  if (length(base) == 0) return(NULL)
  p %>%
    filter(label != "") %>%
    mutate(
      element  = el,
      nominal  = round(mz - base),
      offset_M = round(mz - base, 6)
    ) %>%
    filter(nominal %in% 1:2) %>%
    select(element, level = nominal, isotopologue = label, offset_M, abundance)
})

## All pairwise separations within each nominal level
pairs <- map_dfr(split(ref_raw, ref_raw$level), function(d) {
  if (nrow(d) < 2) return(NULL)
  level_label <- paste0("M+", d$level[1])
  idx <- combn(nrow(d), 2)
  map_dfr(seq_len(ncol(idx)), function(i) {
    ra <- d[idx[1, i], ]; rb <- d[idx[2, i], ]
    tibble(
      level          = level_label,
      `isotopologue A` = paste0(ra$element, ": ", ra$isotopologue),
      `isotopologue B` = paste0(rb$element, ": ", rb$isotopologue),
      `difference (Da)` = round(ra$offset_M - rb$offset_M, 6),
      `abundance A (% of M)` = round(ra$abundance, 4),
      `abundance B (% of M)` = round(rb$abundance, 4)
    )
  })
}) %>%
  arrange(level, abs(`difference (Da)`)) %>%
  mutate(`ppm error` = "")
```

## Search by observed separation

Enter the gap you measured between two fine-structure peaks and a tolerance in
millidaltons (mDa). A Da window makes more physical sense here than ppm — these
separations are just a few tenths of a millidalton, so a 1 mDa window already
covers the instrument's measurement uncertainty regardless of the parent ion's m/z.
The **reference m/z** is used only to compute the **ppm error** display column,
so you can judge the match in the units your instrument reports.

```{r offset-interactive}
#| echo: false
## DataTables core version bundled with DT, for the CDN fallback below
probe_widget    <- datatable(data.frame(x = 1))
DT_CORE_VERSION <- Filter(function(d) d$name == "dt-core",
                           probe_widget$dependencies)[[1]]$version

tagList(
  tags$div(
    style = "display:flex; gap:1em; align-items:end; margin-bottom:0.75em; flex-wrap:wrap;",
    tags$div(
      tags$label("observed gap (Da)", `for` = "iso_off"),
      tags$br(),
      tags$input(id = "iso_off", type = "number", step = "any",
                 class = "form-control", style = "width:12em;")
    ),
    tags$div(
      tags$label("tolerance (mDa)", `for` = "iso_tolmda"),
      tags$br(),
      tags$input(id = "iso_tolmda", type = "number", step = "any", value = "0.05",
                 class = "form-control", style = "width:9em;")
    ),
    tags$div(
      tags$label("reference m/z (Da)", `for` = "iso_refmz"),
      tags$br(),
      tags$input(id = "iso_refmz", type = "number", step = "any", value = "300",
                 class = "form-control", style = "width:10em;")
    ),
    tags$div(
      tags$label("level", `for` = "iso_level"),
      tags$br(),
      tags$select(id = "iso_level", class = "form-control", style = "width:8em;",
        tags$option("All"),
        tags$option("M+1"),
        tags$option("M+2")
      )
    ),
    tags$button("Search", id = "iso_go",    type = "button", class = "btn btn-primary"),
    tags$button("Clear",  id = "iso_clear", type = "button", class = "btn btn-outline-secondary")
  ),
  datatable(pairs, rownames = FALSE, elementId = "iso_table",
            options = list(
              pageLength = 15,
              order = list(list(3, "asc")),
              columnDefs = list(list(targets = 6, searchable = FALSE))
            )),
  tags$script(HTML(sprintf("
    if (typeof $ === 'undefined' || typeof $.fn.dataTable === 'undefined') {
      document.write('<script src=\"https://cdn.datatables.net/%s/js/jquery.dataTables.min.js\"><\\/script>');
    }
    $(document).ready(function() {
      var $tbl = $('#iso_table').find('table').first();
      if (!$tbl.length) return;
      var dt = $tbl.DataTable();
      var target = null, tolDa = null, levelFilter = 'All';

      $.fn.dataTable.ext.search.push(function(settings, data, dataIndex) {
        if (settings.nTable !== $tbl.get(0)) return true;
        if (target === null) return true;
        var val = parseFloat(data[3]);
        return !isNaN(val) && Math.abs(Math.abs(val) - Math.abs(target)) <= tolDa;
      });

      $.fn.dataTable.ext.search.push(function(settings, data, dataIndex) {
        if (settings.nTable !== $tbl.get(0)) return true;
        if (levelFilter === 'All') return true;
        return data[0] === levelFilter;
      });

      function updatePpm() {
        dt.rows().every(function(rowIdx) {
          dt.cell(rowIdx, 6).data('');
        });
        if (target === null) return;
        var ref = parseFloat(document.getElementById('iso_refmz').value);
        if (isNaN(ref) || ref <= 0) ref = 300;
        dt.rows({filter: 'applied'}).every(function(rowIdx) {
          var diff = parseFloat(this.data()[3]);
          var ppm  = ((Math.abs(diff) - Math.abs(target)) / ref * 1e6).toFixed(1);
          dt.cell(rowIdx, 6).data(ppm);
        });
      }

      $('#iso_go').on('click', function() {
        var off    = parseFloat(document.getElementById('iso_off').value);
        var tolMda = parseFloat(document.getElementById('iso_tolmda').value);
        levelFilter = document.getElementById('iso_level').value;
        if (isNaN(off) || isNaN(tolMda)) {
          target = null;
        } else {
          target = off;
          tolDa  = tolMda / 1000;
        }
        dt.draw();
        updatePpm();
        dt.draw(false);
      });

      $('#iso_clear').on('click', function() {
        document.getElementById('iso_off').value   = '';
        document.getElementById('iso_level').value = 'All';
        target = null;
        levelFilter = 'All';
        dt.draw();
        updatePpm();
        dt.draw(false);
      });
    });
  ", DT_CORE_VERSION)))
)
```

The table shows every pair of isotopologues within the same nominal level (M+1 or
M+2). Search by entering the gap you measured between two peaks in the fine
structure.

## How to use this table

1. Zoom into the M+1 or M+2 region of your spectrum and measure the separation
   between any two peaks you can resolve.
2. Enter that gap in the **observed gap** field and a **tolerance in mDa** matching
   your instrument's mass accuracy. Use the **level** dropdown to restrict results
   to M+1 or M+2 if you already know which region you are working in. Click **Search**.
3. Each returned row names the pair of isotopologues whose exact-mass positions are
   that far apart. The **ppm error** column shows how far off the theoretical value
   is relative to the reference m/z — a useful cross-check against your instrument's
   quoted mass accuracy. Check the abundance columns: if one member of the pair is
   far too small to see at your current resolution, rule it out.

For a formula-specific analysis — how large each peak should be, and what resolving
power you need to separate them — see [*Isotope fine structure*](isotope-fine-structure.html).
