---
title: "Reading isotopic fine structure"
subtitle: "Why the M+2 peak is not one peak, and what it takes to see inside it"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    df-print: paged
    fig-format: svg
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svg
vignette: >
  %\VignetteIndexEntry{Reading isotopic fine structure}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(Rdisop)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

## commonMZ's own categorical order (dataviz skill default), used consistently
## for "which element is responsible" throughout every figure in this document.
COL <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
         S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
BLUE <- unname(COL["C"])
```

## The question this document answers

Every ion table in commonMZ — the adducts, the contaminants, the repeating units —
describes a *monoisotopic* mass. But nothing you actually measure is monoisotopic
only: every real peak drags a family of satellites behind it, one nominal mass unit
up, then two, then three. Most of the time you glance at the M+1 and M+2 satellites,
eyeball their height relative to M, and move on — "small M+1, must be a couple of
carbons; big M+2, must be sulfur or chlorine."

That shortcut works because of a coincidence that is worth making explicit: **at
unit mass resolution, M+2 looks like a single peak, but it is actually the sum of
several physically distinct isotopologues that happen to round to the same nominal
mass.** ¹³C¹³C, ³⁴S, ¹⁸O, ³⁷Cl and ⁸¹Br can all land one M+2 satellite depending on
what the molecule contains, and they do not have the same exact mass — each is off
from the true integer by its own small **mass defect**, typically a few
milli-Daltons. At the resolving power routinely available on an Orbitrap, several of
those milli-Dalton gaps are wide enough to *see*: the "M+2 peak" resolves into two
or three peaks that individually confirm which isotopes are actually present.

This document builds that fine structure from first principles with
[Rdisop](https://bioconductor.org/packages/Rdisop/), the same way
[`contaminants_+.tsv`](contaminants_+.tsv)'s harder entries were resolved: every
mass and every natural abundance below comes from Rdisop's own element table, never
typed by hand.

```{r elements}
EL <- initializeElements(c("C", "H", "N", "O", "S", "Cl", "Br", "K", "Na"))
names(EL) <- vapply(EL, `[[`, "", "name")

## Every stable isotope of an element, as Rdisop reports it: k is the nominal
## mass increment from that element's lightest isotope (almost always also its
## most abundant one), `exact` its true mass, `ab` its natural abundance.
isotopes <- function(sym) {
  e <- EL[[sym]]
  tibble(k       = seq_along(e$isotope$mass) - 1,
         massnum = e$mass + k,          # the real isotope, e.g. 13 for 13C
         exact   = e$mass + k + e$isotope$mass,
         ab      = e$isotope$abundance) %>%
    filter(ab > 0)
}

bind_rows(lapply(names(EL), function(s) isotopes(s) %>% mutate(element = s, .before = 1))) %>%
  mutate(exact = round(exact, 5), `abundance %` = round(100 * ab, 4)) %>%
  select(element, isotope = massnum, k, exact, `abundance %`) %>%
  knitr::kable(caption = "Every isotope used in this document, exactly as Rdisop reports it.")
```

Two elements are worth flagging in that table because they don't behave the way
people expect. **Potassium's own +1 satellite is ⁴⁰K**, not some exotic
combination — it just happens to be a trace radioactive isotope (0.012%), so it is
almost never the explanation for an observed M+1. And **oxygen has a stable +1
isotope, ¹⁷O**, sitting at only 0.04% abundance — usually negligible, but it is the
reason a formula-decomposition tool will occasionally offer you an oxygen-based M+1
explanation you didn't expect.

## The coarse pattern everyone learns first

Start with a compound whose isotope pattern needs no explaining: **diclofenac**,
C₁₄H₁₁Cl₂NO₂, an anti-inflammatory drug common enough as a urinary xenobiotic that
it turns up in untargeted metabolomics data. Two chlorines give it a textbook
barcode.

```{r diclofenac-coarse}
#| fig-cap: "Diclofenac's isotope pattern at unit resolution. Two chlorines are unmissable without any fine-structure argument at all."
dic <- getMolecule("C14H11Cl2NO2")
pat <- tibble(iso = seq_along(dic$isotopes[[1]][1, ]) - 1,
              mz  = dic$isotopes[[1]][1, ],
              ab  = dic$isotopes[[1]][2, ]) %>%
  mutate(rel = 100 * ab / max(ab))

ggplot(pat, aes(mz, rel)) +
  geom_segment(aes(xend = mz, y = 0, yend = rel), colour = BLUE, linewidth = 2.2,
               lineend = "round") +
  geom_text(data = filter(pat, rel > 1), aes(label = paste0("M+", iso)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

`getMolecule()$isotopes` is Rdisop doing exactly what every isotope-pattern
calculator does: summing every isotopologue by its **nominal** mass increment and
reporting one number per integer. That is genuinely all you need here — M+2 at 65%
of M is such an overwhelming excess over what carbon alone could produce that no
fine-structure argument is required. Two chlorines announce themselves in a glance.

The interesting cases are the ones where the coarse pattern is *not* dramatic enough
to read at a glance, and the only way to be sure what is in a peak is to look
inside it.

## When the coarse pattern hides the answer

**Methionine**, C₅H₁₁NO₂S, is about as ordinary as a metabolite gets — one of the
twenty proteinogenic amino acids, present in every untargeted metabolomics urine or
plasma run. It also carries one sulfur, and sulfur's own +2 isotope, ³⁴S, sits at
4.25% abundance — two orders of magnitude more abundant than the +1 isotopes of the
CHNO elements. Nothing about a coarse M+1/M+2 barcode makes that obvious the way
diclofenac's two chlorines are obvious.

```{r methionine-coarse}
#| fig-cap: "Methionine's coarse isotope pattern. M+2 is bigger than five carbons alone would ever produce - but the bars alone don't say why."
met <- getMolecule("C5H11NO2S")
patm <- tibble(iso = seq_along(met$isotopes[[1]][1, ]) - 1,
               mz  = met$isotopes[[1]][1, ],
               ab  = met$isotopes[[1]][2, ]) %>%
  mutate(rel = 100 * ab / max(ab))

ggplot(patm, aes(mz, rel)) +
  geom_segment(aes(xend = mz, y = 0, yend = rel), colour = BLUE, linewidth = 2.2,
               lineend = "round") +
  geom_text(data = filter(patm, rel > 1), aes(label = paste0("M+", iso)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Five carbons predict an M+1 of roughly `5 * 1.07% ≈ 5.4%` of M — reasonably close to
what the chart shows, since carbon dominates M+1 for any CHNOS compound this size.
But M+2 at `r round(patm$rel[patm$iso==2],1)`% is the number that should make you
stop: naively, M+2 "ought" to come from picking any two of those five carbons to
both be ¹³C, which combinatorics puts at only
`choose(5,2) * 0.0107^2 * 100 ≈ 0.11%` — **roughly 40 times smaller** than what is
actually observed. Something other than double-¹³C is producing most of that peak,
and the coarse pattern alone cannot tell you what. That gap between the naive
¹³C-only prediction and the observed M+2 height is itself a diagnostic: whenever
M+2 badly overshoots what carbon alone predicts, suspect S, Cl, Br, Si, or a metal
before anything else — which is exactly the reasoning
[`contaminants_+.tsv`](contaminants_+.tsv)'s metal-complex entries were built on.

## Building the fine structure by hand

To go further you have to stop summing by nominal mass and start asking, atom by
atom, *which* isotope substitutions can add up to +1 or +2, and where each one
actually falls in exact mass. For a single molecule with modest atom counts this is
tractable to enumerate directly, and doing it by hand (rather than trusting a
black-box formula) is the whole point — it is the same manual bookkeeping behind
the `first_iso`/`second_iso` reference tables this document's calculators are
descended from, generalised to any formula.

```{r fine-structure-fun}
parse_formula <- function(f) {
  toks <- regmatches(f, gregexpr("[A-Z][a-z]?[0-9]*", f))[[1]]
  el <- sub("^([A-Z][a-z]?)([0-9]*)$", "\\1", toks)
  n  <- sub("^([A-Z][a-z]?)([0-9]*)$", "\\2", toks)
  n[n == ""] <- "1"
  setNames(as.integer(n), el)
}

## Every isotopologue that reaches nominal level `max_level` (1 or 2) by
## substituting one or two atoms, with its exact-mass shift from the
## monoisotopic ion and its abundance relative to that monoisotopic ion.
## Truncating at two substituted atoms is exact for M+1 and M+2 - reaching
## +2 with three or more substitutions is impossible unless an isotope has a
## *negative* nominal increment, which none do.
fine_structure <- function(formula, max_level = 2) {
  counts <- parse_formula(formula)
  iso <- setNames(lapply(names(counts), isotopes), names(counts))

  ## one atom of one element carries a heavier isotope (k = 1 or k = 2 directly,
  ## e.g. 13C, 15N, 2H, 17O, but also 34S, 37Cl, 81Br, 41K, 18O)
  singles <- bind_rows(lapply(names(counts), function(e) {
    it <- iso[[e]]; ex0 <- it$exact[it$k == 0]; ab0 <- it$ab[it$k == 0]
    it %>% filter(k > 0) %>%
      transmute(element = e, k, dmass = exact - ex0,
                rel = counts[[e]] * ab / ab0, label = paste0(massnum, e))
  }))

  ## two DIFFERENT atoms of the SAME element both carry that element's own
  ## +1 isotope (e.g. two different carbons both 13C)
  pair_same <- bind_rows(lapply(names(counts), function(e) {
    it <- iso[[e]]; ex0 <- it$exact[it$k == 0]; ab0 <- it$ab[it$k == 0]
    it %>% filter(k == 1, counts[[e]] >= 2) %>%
      transmute(element = "combo", k = 2 * k, dmass = 2 * (exact - ex0),
                rel = choose(counts[[e]], 2) * (ab / ab0)^2,
                label = paste(rep(paste0(massnum, e), 2), collapse = ", "))
  }))

  ## one atom from each of two DIFFERENT elements each carry a +1 isotope
  ## (e.g. one 13C and one 15N)
  ones <- singles %>% filter(k == 1)
  pair_diff <- tibble()
  if (nrow(ones) >= 2) {
    ix <- combn(seq_len(nrow(ones)), 2, simplify = FALSE)
    pair_diff <- bind_rows(lapply(ix, function(p) {
      a <- ones[p[1], ]; b <- ones[p[2], ]
      tibble(element = "combo", k = 2, dmass = a$dmass + b$dmass,
             rel = a$rel * b$rel, label = paste(a$label, b$label, sep = ", "))
    }))
  }

  bind_rows(tibble(element = "M", k = 0, dmass = 0, rel = 1, label = "M (monoisotopic)"),
            singles, pair_same, pair_diff) %>%
    filter(k <= max_level) %>%
    arrange(k, desc(rel))
}
```

Before trusting this on anything new, check it against ground truth: Rdisop's own
`getMolecule()$isotopes` sums every isotopologue by brute-force combinatorics with
no truncation at all. If the hand-built decomposition above is summed back up by
nominal level, it should reproduce Rdisop's numbers exactly.

```{r verify}
verify <- function(formula) {
  fs <- fine_structure(formula) %>% group_by(k) %>% summarise(mine = 100 * sum(rel), .groups = "drop")
  mol <- getMolecule(formula)
  truth <- tibble(k = seq_along(mol$isotopes[[1]][1, ]) - 1,
                   rdisop = 100 * mol$isotopes[[1]][2, ] / mol$isotopes[[1]][2, 1]) %>%
    filter(k %in% fs$k)
  left_join(fs, truth, by = "k") %>% mutate(across(c(mine, rdisop), ~round(.x, 4)))
}
bind_rows(verify("C5H11NO2S") %>% mutate(formula = "methionine", .before = 1),
          verify("C14H11Cl2NO2") %>% mutate(formula = "diclofenac", .before = 1)) %>%
  knitr::kable(caption = "Hand-built fine structure, summed by nominal level, against Rdisop's own combinatorics. They agree to four decimal places.")
```

They match. That agreement is the whole justification for looking inside the sum
below.

## What is actually inside methionine's M+2

```{r met-m1}
#| fig-cap: "Methionine, M+1: five candidate isotope substitutions, spread across 9 mDa. Carbon dominates, as it does for almost every organic ion."
fs_met <- fine_structure("C5H11NO2S")
m0_met <- getMolecule("C5H11NO2S")$exactmass

plot_level <- function(fs, m0, level, title) {
  d <- fs %>% filter(k == level) %>%
    mutate(mz = m0 + dmass, mDa = round((dmass - level) * 1000, 2),
           col = ifelse(element %in% names(COL), element, "combo"))
  ggplot(d, aes(mDa, rel, colour = col)) +
    geom_segment(aes(xend = mDa, y = 0, yend = rel), linewidth = 2.2, lineend = "round") +
    geom_text(data = slice_max(d, rel, n = min(6, nrow(d))),
              aes(label = label), angle = 90, hjust = -0.15, size = 3.1, colour = "grey20") +
    scale_colour_manual(values = COL, guide = "none") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .35))) +
    labs(x = "mDa from the nominal integer mass", y = "% of M (monoisotopic)", title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
plot_level(fs_met, m0_met, 1, "Methionine M+1")
```

```{r met-m2}
#| fig-cap: "Methionine, M+2: 34S alone outweighs every carbon-only combination roughly 40-fold, exactly matching the excess spotted in the coarse pattern above."
plot_level(fs_met, m0_met, 2, "Methionine M+2")
```

Two things the coarse bar chart could never show are visible here. First,
**sulfur's own +2 isotope physically sits apart from every carbon-based
explanation** — `13C, 13C` (two different carbons, both ¹³C) is
`r round(fs_met$dmass[fs_met$label=="13C, 13C"]*1000-2000,2)`
mDa away from `34S`, a large enough gap that they are two genuinely different masses,
not two names for the same peak. Second, the height difference *is* the sulfur
signal: `34S` towers over every combination of carbons, nitrogens and oxygens that
could also reach +2, which is the fine-structure version of the 40-fold excess
already spotted in the coarse pattern — now with a specific competing hypothesis
(¹³C¹³C) sitting at a specific, different mass, instead of a vague "something else."

## Would your instrument actually separate them?

A mass difference on paper is not the same as two resolved peaks on a chromatogram.
Whether `34S` and `13C, 13C` show up as one blob or two depends entirely on the
resolving power available at that m/z. Orbitraps are quoted at a reference m/z
(typically 200) and lose resolving power roughly as `1/sqrt(m/z)` away from it:

```{r resolving-power}
R_at_mz  <- function(mz, R_ref, mz_ref = 200) R_ref * sqrt(mz_ref / mz)
fwhm_at  <- function(mz, R_ref, mz_ref = 200) mz / R_at_mz(mz, R_ref, mz_ref)

## every pairwise gap within one nominal level, and the resolving power (at
## the reference m/z used above) that would put a valley between the peaks.
## min_abundance drops candidates too faint to ever matter in practice (e.g.
## two deuteriums at once) so the pairwise table stays a short, readable list
## of the comparisons an analyst could actually be faced with.
separations <- function(fs, m0, level, R_ref = 120000, mz_ref = 200, min_abundance = 0.0001) {
  d <- fs %>% filter(k == level, rel >= min_abundance) %>% mutate(mz = m0 + dmass)
  if (nrow(d) < 2) return(tibble())
  ix <- combn(seq_len(nrow(d)), 2)
  bind_rows(lapply(seq_len(ncol(ix)), function(i) {
    ra <- d[ix[1, i], ]; rb <- d[ix[2, i], ]
    mDa  <- abs(ra$mz - rb$mz) * 1000
    Rreq <- (ra$mz + rb$mz) / 2 / abs(ra$mz - rb$mz)          # R needed AT that m/z
    Rref_req <- Rreq / sqrt(mz_ref / mean(c(ra$mz, rb$mz)))   # translated to the R_ref convention
    tibble(a = ra$label, b = rb$label, abundance_a_pct = round(ra$rel, 3),
           abundance_b_pct = round(rb$rel, 3), separation_mDa = round(mDa, 2),
           R_needed_here = round(Rreq), R_needed_at200 = round(Rref_req))
  })) %>% arrange(R_needed_at200)
}

sep_met2 <- separations(fs_met, m0_met, 2)
sep_met2 %>%
  knitr::kable(col.names = c("a", "b", "abundance a %", "abundance b %",
                              "separation (mDa)", "R needed here", "R needed @200"),
               caption = "Every pair of methionine M+2 candidates above 0.01% abundance, and the resolving power (quoted at m/z 200, Orbitrap convention) that would separate them.")
```

`34S` vs `13C, 13C` needs only R ≈ 14,000 at m/z 200 — trivial for any current Orbitrap,
even the fastest, lowest-resolution methods. The pair that is genuinely hard is
inside methionine's M+1: distinguishing `13C` from `17O` needs roughly
`r { s <- separations(fs_met, m0_met, 1); hit <- (s$a == "13C" & s$b == "17O") | (s$a == "17O" & s$b == "13C"); format(round(s$R_needed_at200[hit], -3), big.mark = ",", scientific = FALSE) }`
— a resolving power few instruments carry. That contrast is the actual lesson: some
fine-structure questions are free, others are not, and the only way to know which is
to compute the gap.

Numbers on a table are one thing; seeing whether two peaks actually merge is more
convincing. Simulate the region as it would really look, summing a Gaussian peak of
the right width for each candidate at three realistic resolving powers:

```{r sim-profile}
#| fig-cap: "Methionine's M+2 region simulated at three Orbitrap resolving powers (quoted @ m/z 200). Vertical ticks mark where each candidate actually sits. At 15,000 the shoulder is invisible; by 60,000 it is unambiguous."
simulate_profile <- function(fs, m0, level, R_ref, mz_ref = 200, n = 3000, pad_mDa = 15) {
  d <- fs %>% filter(k == level) %>% mutate(mz = m0 + dmass)
  grid <- seq(min(d$mz) - pad_mDa / 1000, max(d$mz) + pad_mDa / 1000, length.out = n)
  y <- Reduce(`+`, lapply(seq_len(nrow(d)), function(i) {
    sigma <- fwhm_at(d$mz[i], R_ref, mz_ref) / (2 * sqrt(2 * log(2)))
    d$rel[i] * exp(-(grid - d$mz[i])^2 / (2 * sigma^2))
  }))
  tibble(mz = grid, y = y)
}

Rs <- c(15000, 60000, 120000)
R_label <- function(r) sprintf("R = %s @200", format(r, big.mark = ",", trim = TRUE))
prof <- bind_rows(lapply(Rs, function(r)
  simulate_profile(fs_met, m0_met, 2, r) %>% mutate(R = R_label(r))))
ticks <- fine_structure("C5H11NO2S") %>% filter(k == 2, rel > 0.02) %>% mutate(mz = m0_met + dmass)

ggplot(prof, aes(mz, y)) +
  geom_area(fill = BLUE, alpha = .25) +
  geom_line(colour = BLUE, linewidth = .8) +
  geom_rug(data = ticks, aes(x = mz), inherit.aes = FALSE, sides = "b", colour = "grey30", linewidth = .6) +
  facet_wrap(~ factor(R, levels = vapply(Rs, R_label, "")), ncol = 1, scales = "free_y") +
  labs(x = "m/z", y = "simulated intensity") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(hjust = 0))
```

## Diclofenac's fine structure, for contrast

Diclofenac didn't need any of this — its M+2 excess was obvious at unit resolution.
Its fine structure is still there, though, and it makes the contrast explicit:

```{r dic-m2}
#| fig-cap: "Diclofenac M+2: two chlorines dominate so completely that the tiny 13C,13C shoulder barely registers, even though it sits at a perfectly resolvable distance."
fs_dic <- fine_structure("C14H11Cl2NO2")
m0_dic <- getMolecule("C14H11Cl2NO2")$exactmass
plot_level(fs_dic, m0_dic, 2, "Diclofenac M+2")
```

`37Cl` is
`r { sd <- separations(fs_dic, m0_dic, 2); hit <- (sd$a=="37Cl" & sd$b=="13C, 13C") | (sd$a=="13C, 13C" & sd$b=="37Cl"); sd$separation_mDa[hit] }`
mDa from `13C, 13C` and roughly 60 times taller — a case where the fine structure
confirms what the coarse pattern already screamed, rather than revealing something
the coarse pattern hid. Methionine and diclofenac are the two ends of the same
spectrum: sometimes the isotope pattern tells you everything at unit resolution,
and sometimes the honest answer is sitting a few milli-Daltons away from a much
more obvious but wrong explanation, and the only way to tell the two apart is to
do the arithmetic above.

## Reading a real fine-structure peak: a checklist

1. **Start from the coarse pattern.** Compare the observed M+1 and M+2 heights to
   what carbon count alone predicts (`nC * 1.07%` for M+1;
   `choose(nC,2) * 0.0107^2 * 100%` for M+2). A large excess is the first signal
   that something heavier than carbon is present.
2. **Enumerate the candidates** with `fine_structure()` (or by hand for a small
   formula) to see which specific isotopologues could produce that excess, and how
   far apart they sit in exact mass.
3. **Check the gap against your resolving power** with `separations()` before
   concluding a peak is "clean" — a shoulder that should be there but isn't
   resolved is not evidence it's absent, only evidence you didn't have the
   resolving power to see it.
4. **Verify against Rdisop's own combinatorics** (as in `verify()` above) whenever
   the formula is complex enough that hand enumeration feels risky — never trust a
   hand-built decomposition you haven't checked against ground truth.

None of the functions above are exported by commonMZ; they exist to make this
document's arithmetic reproducible and auditable, in the same spirit as every
number in `contaminants_+.tsv`.
