---
title: "Isotope fine structure"
subtitle: "Why M+1 and M+2 each contain multiple peaks, and how to simulate what your instrument shows"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    fig-format: svg
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Isotope fine structure}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(dplyr)
library(purrr)
library(ggplot2)
library(plotly)

BLUE <- "#2a78d6"
COL  <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
         S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
```

Isotope fine structure is what happens when a nominal M+1 or M+2 level contains
several isotopologues at slightly different exact masses. At **M+1**, every peak
comes from substituting exactly one atom with its heavy isotope — ¹³C, ²H, and ¹⁵N
are the most prominent contributors (oxygen and sulfur are better known for M+2:
¹⁸O and ³⁴S are five times more abundant than ¹⁷O and ³³S). At **M+2**, single
substitutions of two-neutron isotopes (³⁴S, ³⁷Cl, ¹⁸O…) appear alongside
combinations of two M+1 isotopes. Whether your instrument resolves any of these as separate peaks
depends on resolving power. This article explains the positions, walks through a
concrete example, and builds the tools to simulate what a real spectrum would show.

## M+1: one atom is a heavier isotope, but you get several exact masses

If you add a neutron to a nucleus, the mass increase is not exactly the free neutron
mass (1.008665 Da). When the neutron binds, the nucleus releases some energy — and
by E = mc², that released energy corresponds to a small mass deficit. The binding
energy gain differs for every element, so each isotope has its own exact mass shift.
All of the following contribute to the M+1 level, but they sit at genuinely different
m/z values:

| Isotopologue | Shift from monoisotopic (Da) | Offset from ¹³C (mDa) |
|---|---:|---:|
| ¹³C | +1.003355 | 0 |
| ¹⁷O | +1.004220 | +0.9 |
| ²H  | +1.006277 | +2.9 |
| ³³S | +0.999388 | −3.9 |
| ¹⁵N | +0.997035 | −6.3 |

The 9.2 mDa spread between ¹⁵N and ²H is larger than the mass accuracy of any
modern high-resolution instrument. Whether your spectrum resolves them as separate
peaks depends on resolving power and the compound's elemental composition; the
worked example below shows how to calculate that.

**Why ¹³C tends to dominate M+1.** The relative height of each peak is proportional
to the isotope's natural abundance multiplied by the number of atoms of that element
in the molecule. ¹³C has 1.11% abundance per carbon, and organic molecules often
contain many carbons — so the total ¹³C contribution (roughly *n* × 1.11%) quickly
outweighs ²H (0.015% per H), ¹⁵N (0.37% per N), and the trace M+1 isotopes of
O and S. A molecule with five carbons already has a ~5.5% chance of carrying one
¹³C, far above any single ¹⁵N or ²H contribution.

## M+2: combinations and heavier isotopes

At M+2, two types of peak appear: two-neutron heavy isotopes (³⁴S, ³⁷Cl, ³⁸Ar…)
and combinations of two M+1 isotopes. They sit at different exact masses:

| Isotopologue | Shift from monoisotopic (Da) | Offset from ¹³C,¹³C (mDa) |
|---|---:|---:|
| ¹³C + ¹³C | +2.006710 | 0 |
| ¹⁸O | +2.004220 | −2.5 |
| ³⁷Cl | +1.997051 | −9.7 |
| ³⁴S | +1.995796 | −10.9 |

When a molecule carries several chlorines or one sulfur, these two-neutron isotopes
can completely dominate the M+2 level. **Diclofenac** (C₁₄H₁₁Cl₂NO₂) is an
extreme case: two chlorines at 24.2% each push M+2 to ~65% of M, an excess so
large that no fine-structure measurement is needed to recognise it.

```{r diclofenac-coarse}
#| fig-cap: "Diclofenac's isotope pattern at unit resolution. Two chlorines make M+2 unmissable without any fine-structure argument."
pat_dic  <- isotope_fine_pattern("C14H11Cl2NO2")
base_dic <- pat_dic %>% filter(label == "") %>% pull(mz)

coarse_dic <- pat_dic %>%
  mutate(nominal = round(mz - base_dic)) %>%
  group_by(nominal) %>%
  summarise(mz = base_dic + first(nominal), abundance = sum(abundance), .groups = "drop")

ggplot(coarse_dic, aes(mz, abundance)) +
  geom_segment(aes(xend = mz, y = 0, yend = abundance),
               colour = BLUE, linewidth = 2.2, lineend = "round") +
  geom_text(data = filter(coarse_dic, abundance > 1),
            aes(label = paste0("M+", nominal)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z (Da)", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Running example: methionine

**Methionine** (C₅H₁₁NO₂S) is one of the twenty proteinogenic amino acids, present
in every untargeted metabolomics run. It carries one sulfur, whose ³⁴S isotope
sits at 4.25% natural abundance. Nothing about a coarse M+1/M+2 barcode makes that
obvious: the bars look ordinary until you notice M+2 far exceeds what five carbons
alone could explain.

Change `formula` here to run the same workflow for your own candidate:

```{r formula}
formula <- "C5H11NO2S"   # methionine -- swap to your own candidate

pat     <- isotope_fine_pattern(formula)
base_mz <- pat %>% filter(label == "") %>% pull(mz)
coarse  <- pat %>%
  mutate(nominal = round(mz - base_mz)) %>%
  group_by(nominal) %>%
  summarise(mz = base_mz + first(nominal), abundance = sum(abundance), .groups = "drop")
```

```{r methionine-coarse}
#| fig-cap: "Methionine's coarse isotope pattern. M+2 far exceeds what five carbons alone would produce, but the bars don't say why."
ggplot(coarse, aes(mz, abundance)) +
  geom_segment(aes(xend = mz, y = 0, yend = abundance),
               colour = BLUE, linewidth = 2.2, lineend = "round") +
  geom_text(data = filter(coarse, abundance > 1),
            aes(label = paste0("M+", nominal)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15))) +
  labs(x = "m/z (Da)", y = "relative abundance (%)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Five carbons predict M+2 from double-¹³C at
`choose(5,2) * 0.0107^2 * 100 ≈ 0.11%`, **roughly 40 times smaller** than the
`r round(coarse %>% filter(nominal==2) %>% pull(abundance), 1)`% actually observed.
Whenever M+2 badly overshoots what carbon alone predicts, suspect S, Cl, Br, Si, or
a metal — the fine structure below names the specific isotopologue.

## Step 1: build the fine-structure table

`isotope_fine_pattern()` reports each isotopologue separately, at its own exact
mass, labelled by which isotopes produced it:

```{r fine-table}
pat %>% arrange(mz) %>%
  mutate(mz = round(mz, 5), abundance = round(abundance, 4),
         label = ifelse(label == "", "M (monoisotopic)", label)) %>%
  knitr::kable(caption = paste("Every isotopologue of", formula, "above the default 0.01% threshold."))
```

## Step 2: zoom into one nominal level

On the m/z axis, M+1 and M+2 each split into several distinct peaks:

```{r met-m1}
#| fig-cap: "M+1: five candidate isotope substitutions, spread across 0.009 Da. Carbon dominates, as it does for almost every organic ion."
plot_level <- function(pattern, level, title) {
  base <- pattern %>% filter(label == "") %>% pull(mz)
  d <- pattern %>%
    mutate(nominal = round(mz - base)) %>%
    filter(nominal == level) %>%
    mutate(element = sub(",.*", "", label),
           element = sub("^[0-9]+", "", element),
           col = ifelse(label == "" | grepl(",", label), "combo", element),
           col = ifelse(col %in% names(COL), col, "combo"))
  ggplot(d, aes(mz, abundance, colour = col)) +
    geom_segment(aes(xend = mz, y = 0, yend = abundance),
                 linewidth = 2.2, lineend = "round") +
    geom_text(data = slice_max(d, abundance, n = min(6, nrow(d))),
              aes(label = label),
              angle = 90, hjust = -0.15, size = 3.1, colour = "grey20") +
    scale_colour_manual(values = COL, guide = "none") +
    scale_x_continuous(labels = scales::label_number(accuracy = 0.0001)) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .35))) +
    labs(x = "m/z (Da)", y = "% of M (monoisotopic)", title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
plot_level(pat, 1, paste(formula, "M+1"))
```

At M+1, ¹³C dominates (five carbons × 1.07%) with ¹⁵N and ³³S present but roughly
20-fold smaller. Their mass offsets from ¹³C are −6.3 mDa and −3.9 mDa
respectively — both resolvable in principle on a high-resolution instrument,
though the abundance difference means ¹³C will always be the tallest peak.

```{r met-m2}
#| fig-cap: "M+2: 34S alone outweighs every carbon-only combination roughly 40-fold and sits at a measurably different mass."
plot_level(pat, 2, paste(formula, "M+2"))
```

At M+2, the same two things the coarse bar chart could not show are now visible.
First, **³⁴S physically sits apart from every carbon-based explanation**: `13C, 13C`
sits
`r { d2 <- pat %>% mutate(nominal = round(mz - base_mz)) %>% filter(nominal == 2); round(abs((d2 %>% filter(label=="13C, 13C") %>% pull(mz)) - (d2 %>% filter(label=="34S") %>% pull(mz))), 5) }`
Da away from `34S`.
Second, the height difference *is* the sulfur signal: `34S` towers over every
carbon/nitrogen/oxygen combination that could also reach +2 — the fine-structure
version of the 40-fold excess already spotted in the coarse pattern, now with a
specific competing hypothesis (¹³C¹³C) at a specific, different mass.

## Step 3: would your instrument resolve the candidates?

For every pair of isotopologues at a nominal level, the table below reports the
Orbitrap resolving power (quoted at m/z 200) at which the two Gaussian peaks would
show a ~50% valley between them, enough for a centroid algorithm to find two
separate peaks. Two peaks of equal height separated by 1.5 × FWHM produce that
valley; translating to the R@200 Orbitrap convention scales as `1/sqrt(m/z)`:

```{r resolving-power}
R_at_mz <- function(mz, R_ref, mz_ref = 200) R_ref * sqrt(mz_ref / mz)

separations <- function(pattern, level, mz_ref = 200, min_abundance = 0.01) {
  base <- pattern %>% filter(label == "") %>% pull(mz)
  d <- pattern %>%
    mutate(nominal = round(mz - base)) %>%
    filter(nominal == level, abundance >= min_abundance) %>%
    mutate(label = ifelse(label == "", "M", label))
  if (nrow(d) < 2) return(tibble())
  ix <- combn(seq_len(nrow(d)), 2)
  map_dfr(seq_len(ncol(ix)), function(i) {
    ra <- d[ix[1, i], ]; rb <- d[ix[2, i], ]
    sep_da   <- abs(ra$mz - rb$mz)
    m_avg    <- (ra$mz + rb$mz) / 2
    Rreq     <- 1.5 * m_avg / sep_da
    Rref_req <- Rreq / sqrt(mz_ref / m_avg)
    tibble(a = ra$label, b = rb$label,
           abundance_a_pct = round(ra$abundance, 3),
           abundance_b_pct = round(rb$abundance, 3),
           separation_Da = round(sep_da, 5),
           R_at_mz = round(Rreq), R_at200 = round(Rref_req))
  }) %>% arrange(R_at200)
}

ion_mz <- base_mz + 1.007276   # [M+H]+ in positive mode

sep2 <- separations(pat, 2)
sep2 %>%
  knitr::kable(col.names = c("a", "b", "abundance a %", "abundance b %",
                              "separation (Da)", paste0("R at m/z ", round(ion_mz)), "R @200"),
               caption = "Every pair of M+2 candidates above 0.01% abundance. R@200 is the Orbitrap resolving power at which the two peaks show a ~50% valley.")
```

```{r sep-text, results = "asis"}
best <- sep2 %>% slice_min(separation_Da, n = 1) %>% slice(1)
cat(sprintf("The closest pair (`%s` vs `%s`, %.4f Da apart) needs R ~ %s @200 to resolve as two peaks.",
            best$a, best$b, best$separation_Da, format(best$R_at200, big.mark = ",")))
```

## Step 4: simulate the profile

`isotope_profile()` turns the fine-structure table into simulated peak shapes at any
resolving power. Pass it the *actual* resolving power at the ion's m/z (here [M+H]⁺ at m/z `r round(ion_mz)`),
converting from the instrument's `R @ 200` specification with `R_at_mz()`:

```{r sim-profile}
Rs <- c(15000, 60000, 120000)
R_label <- function(r) sprintf("R = %s @200", format(r, big.mark = ",", trim = TRUE))
pat2 <- pat %>% mutate(nominal = round(mz - base_mz)) %>% filter(nominal == 2)

prof <- map_dfr(Rs, function(r)
  isotope_profile(pat2, resolution = R_at_mz(mean(pat2$mz), r)) %>%
    mutate(R = R_label(r)))

p <- ggplot(prof, aes(mz, intensity)) +
  geom_area(fill = BLUE, alpha = .25) +
  geom_line(colour = BLUE, linewidth = .8) +
  geom_rug(data = pat2, aes(x = mz),
           inherit.aes = FALSE, sides = "b", colour = "grey30", linewidth = .6) +
  facet_wrap(~ factor(R, levels = map_chr(Rs, R_label)), ncol = 1, scales = "free_y") +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.0001)) +
  labs(x = "m/z (Da)", y = "simulated intensity (% of M)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(hjust = 0))
ggplotly(p, tooltip = c("x", "y"))
```

At 15,000 the ³⁴S peak is a single, slightly lopsided blob; the ¹³C¹³C shoulder is
in there but invisible. By 60,000 it is an unmistakable second peak. Fine structure
is not a yes/no property of a molecule; it is a question you can only answer once
you know both *what* could be there and *whether your instrument could show it to you*.

## Reference: first and second isotope fine structure

The charts below position every M+1 (first isotope) and M+2 (second isotope)
isotopologue by its mass offset from the **carbon reference peak** (¹³C for M+1;
¹³C,¹³C for M+2), normalized so the tallest bar reaches 1. This shows at a glance
which non-carbon elements contribute and how far they sit from the carbon signal on
the m/z axis.

```{r iso-ref-helper}
#| include: false
iso_ref_data <- function(pattern, level, ref_label) {
  base <- pattern %>% filter(label == "") %>% pull(mz)
  d <- pattern %>%
    mutate(nominal = round(mz - base)) %>%
    filter(nominal == level)
  ref_mz <- d %>% filter(label == ref_label) %>% pull(mz)
  if (length(ref_mz) == 0) ref_mz <- d %>% slice_max(abundance, n = 1) %>% pull(mz)
  d %>%
    mutate(offset    = round(mz - ref_mz, 5),
           rel_abund = abundance / max(abundance)) %>%
    arrange(offset)
}

plot_iso_ref <- function(d, ref_label, level_label) {
  p <- ggplot(d, aes(offset, rel_abund,
                     text = paste0(label, "\noffset: ", offset, " Da\n",
                                   round(abundance, 4), "% of M"))) +
    geom_segment(aes(xend = offset, y = 0, yend = rel_abund),
                 linewidth = 1.5, lineend = "round") +
    geom_text(data = filter(d, rel_abund > 0.005),
              aes(label = label),
              angle = 90, hjust = -0.1, size = 3, colour = "grey20") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .4))) +
    labs(x = paste0("mass offset from ", ref_label, " (Da)"),
         y = paste0("relative abundance within M+", level_label)) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
  ggplotly(p, tooltip = "text")
}
```

### First isotope (M+1)

```{r first-iso-chart}
#| fig-cap: "M+1 fine structure: each bar is one isotopologue; x-axis is mass offset from ¹³C. The carbon peak sits at zero by definition. Hover for exact values."
m1 <- iso_ref_data(pat, 1, "13C")
plot_iso_ref(m1, "¹³C", 1)
```

```{r first-iso-table}
m1 %>%
  select(label, offset, abundance) %>%
  rename(isotopologue              = label,
         `offset from 13C (Da)`   = offset,
         `abundance (% of M)`     = abundance) %>%
  knitr::kable(digits = 5,
               caption = paste("M+1 isotopologues of", formula, "by mass offset from ¹³C."))
```

```{r first-iso-diff}
separations(pat, 1) %>%
  knitr::kable(col.names = c("a", "b", "abundance a %", "abundance b %",
                              "separation (Da)", paste0("R at m/z ", round(ion_mz)), "R @200"),
               caption = "Pairwise separations between M+1 candidates above 0.01% abundance.")
```

### Second isotope (M+2)

```{r second-iso-chart}
#| fig-cap: "M+2 fine structure: x-axis is mass offset from ¹³C,¹³C. Elements with lower exact mass than two carbons (S, Cl, Br...) appear at negative offsets. Hover for exact values."
m2 <- iso_ref_data(pat, 2, "13C, 13C")
plot_iso_ref(m2, "¹³C,¹³C", 2)
```

```{r second-iso-table}
m2 %>%
  select(label, offset, abundance) %>%
  rename(isotopologue                   = label,
         `offset from 13C,13C (Da)`    = offset,
         `abundance (% of M)`          = abundance) %>%
  knitr::kable(digits = 5,
               caption = paste("M+2 isotopologues of", formula, "by mass offset from ¹³C,¹³C."))
```

```{r second-iso-diff}
separations(pat, 2) %>%
  knitr::kable(col.names = c("a", "b", "abundance a %", "abundance b %",
                              "separation (Da)", paste0("R at m/z ", round(ion_mz)), "R @200"),
               caption = "Pairwise separations between M+2 candidates above 0.01% abundance.")
```

## Checklist

1. **Build the pattern** with `isotope_fine_pattern(formula)` and check the coarse
   sum against what carbon count alone predicts (`nC * 1.07%` for M+1;
   `choose(nC,2) * 0.0107^2 * 100%` for M+2). A large excess flags S, Cl, Br, Si,
   or a metal.
2. **Zoom into the nominal level** that shows the excess to see which isotopologues
   are present and how far apart they sit in exact mass (in Da, directly on your
   spectrum's axis).
3. **Check the gap against your resolving power** with `separations()` before
   concluding a peak is clean: an unresolved shoulder is not evidence the candidate
   is absent, only that the instrument couldn't show it.
4. **Simulate the profile** with `isotope_profile()` at your instrument's actual
   resolving power (converted from its `R @ reference m/z` spec with `R_at_mz()`)
   to see, rather than infer, whether the candidates merge.

`plot_level()` and `separations()` are not exported by commonMZ; only
`isotope_fine_pattern()` and `isotope_profile()` are. Both helpers are short enough
to copy into your own analysis.

If you measured a gap between two resolved fine-structure peaks and want to identify
which pair of isotopologues produced it, see
[*Looking up an isotopologue offset*](isotope-offset-lookup.html).
