---
title: "Looking up an unexplained mass difference"
subtitle: "Problem: I have a delta between two peaks and no formula yet. What could it be?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    fig-format: svg

execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Looking up an unexplained mass difference}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(ggplot2)
library(dplyr)
library(DT)
library(RefManageR)

COL <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
        S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
```

```{r fix-dt-core}
#| echo: false
htmltools::includeScript(
  system.file("htmlwidgets/lib/datatables/js/jquery.dataTables.min.js", package = "DT")
)
```

You have two peaks in a spectrum and a delta between them, and the question is "what
family of thing produces a gap this size?" An in-source fragment loses a neutral
(water, CO₂, a whole side chain). A homologous series steps by a repeating unit
(CH₂, C₂H₄O, a whole PEG monomer). An adduct swaps one attached ion for another
(Na⁺ for K⁺, formate for acetate). All three are catalogued in commonMZ already:
`adducts_fragments` and `repeating_units_pos`/`repeating_units_neg`, merged by
`mz_diff_table()` into one searchable reference. Once you have a candidate formula,
[*Isotope fine structure*](isotope-fine-structure.html) takes over.

```{r diff-table-build}
diffs <- mz_diff_table("both")
nrow(diffs)
```

Isotope spacings are deliberately not in this table; they depend on which formula
produced the peaks and belong to
[the isotope fine-structure tool](isotope-fine-structure.html) instead.

## The calculator: search by a difference and a ppm tolerance

This is `mz_diff_lookup()`'s job for a single value called from R. Tolerance is in
ppm of the difference by default (the usual way instrument accuracy is quoted),
though note that is *not* the same as ppm of either peak's own m/z; see the
function's documentation (`?mz_diff_lookup`) for why that distinction matters and
when to widen the tolerance or pass an absolute `unit = "Da"` window instead.

```{r lookup-example}
mz_diff_lookup(18.0106, tol = 100) %>%   # water, 100 ppm of the delta
  mutate(error_Da = signif(error_Da, 2), error_ppm = round(error_ppm, 1)) %>%
  DT::datatable(rownames = FALSE,
                options = list(pageLength = 10, dom = "t"))
```

And a delta genuinely ambiguous without more context: several catalogued
differences sit within a few ppm of each other, so the delta alone does not
decide; you still need to ask whether the sample plausibly contains an alkane
chain, an acrylamide adduct, or neither:

```{r lookup-example-2}
mz_diff_lookup(14.0157, tol = 50) %>%
  mutate(error_Da = signif(error_Da, 2), error_ppm = round(error_ppm, 1)) %>%
  DT::datatable(rownames = FALSE,
                options = list(pageLength = 10, dom = "t"))
```

## Working interactively

For browsing without writing R code, type a difference and a ppm tolerance and
click *Search*. The table's own *Search:* box (top right) still works as usual
for browsing by category, mode or origin text.

```{r diff-interactive}
#| echo: false
htmltools::tagList(
  htmltools::tags$div(
    style = "display:flex; gap:1em; align-items:end; margin-bottom:0.75em; flex-wrap:wrap;",
    htmltools::tags$div(
      htmltools::tags$label("m/z difference (Da)", `for` = "mdl_mz"),
      htmltools::tags$br(),
      htmltools::tags$input(id = "mdl_mz", type = "number", step = "any",
                            class = "form-control", style = "width:12em;")
    ),
    htmltools::tags$div(
      htmltools::tags$label("tolerance (ppm)", `for` = "mdl_ppm"),
      htmltools::tags$br(),
      htmltools::tags$input(id = "mdl_ppm", type = "number", step = "any", value = "50",
                            class = "form-control", style = "width:10em;")
    ),
    htmltools::tags$button("Search", id = "mdl_go", type = "button", class = "btn btn-primary"),
    htmltools::tags$button("Clear", id = "mdl_clear", type = "button", class = "btn btn-outline-secondary")
  ),
  datatable(diffs, rownames = FALSE, elementId = "mdl_table",
            colnames = c("mz_diff (Da)", "category", "mode", "origin", "reference"),
            options = list(pageLength = 8, order = list(list(0, "asc")))),
  htmltools::includeScript("mass-diff-search.js")
)
```

The search only re-filters when you click *Search* (not on every keystroke), so
typing a value never triggers a redraw mid-edit; click once you've entered both
numbers, or press *Clear* to go back to the full table.

## Checklist

1. **Search first, don't guess one hypothesis at a time.** `mz_diff_lookup()` (or
   the interactive table above) checks your delta against every known
   adduct/fragment/repeating-unit difference in commonMZ at once.
2. **A delta rarely has only one explanation.** When several catalogued
   differences land within tolerance of each other, use what you know about the
   sample (the matrix, the method, what's chemically plausible) to break the
   tie, the same way the mobile-phase composition was what actually resolved the
   ambiguous formulas in `contaminants_+.tsv`'s metal-complex entries.
3. **Mind the tolerance convention.** ppm of a small delta is a much tighter
   window than ppm of a large one; if nothing matches, try a wider tolerance or an
   absolute Da window before concluding the delta is uncatalogued.
4. **If you do have (or can guess) a formula**, switch to
   [*Isotope fine structure*](isotope-fine-structure.html):
   an isotope satellite is a different kind of delta than this table covers.

## Glossary

### Ion types

| Notation | Meaning |
|---|---|
| `f+` | fragment ion |
| `[f+H]+` | protonated fragment ion (e.g. in-source fragmentation) |
| `[M+H]+` | protonated molecular ion (pseudomolecular ion) |
| `[M+Na]+` | sodiated molecular ion |
| `[M+K]+` | potassiated molecular ion |
| `[2M+H]+`, `[3M+H]+` | protonated dimer, trimer, etc. |
| `[AnBm+H]+` | protonated ion of a complex with *n* A and *m* B subunits |

### Abbreviations used in the *origin* column

| Abbreviation | Meaning |
|---|---|
| 4-HCCA | α-cyano-4-hydroxycinnamic acid — common MALDI matrix |
| 2,5-DHB | 2,5-dihydroxybenzoic acid — common MALDI matrix |
| MeCN, ACN | acetonitrile (solvent) |
| MeOH | methanol (solvent) |
| MeNO₂ | nitromethane (solvent) |
| HABA | 2-(4-hydroxyphenylazo)benzoic acid — MALDI matrix |
| SA | sinapic / sinapinic acid — common MALDI matrix |
| PEG | polyethylene glycol; repeat unit –[O–CH₂–CH₂]–, 44 Da |
| PPG | polypropylene glycol; repeat unit –[O–C(CH₃)H–CH₂]–, 58 Da |
| XaaCcamXaa | carbamidomethylated cysteine residue (+57 Da) |
| XaaMoxXaa | singly oxidised methionine residue (+16 Da) |

### References for the *reference* column

```{r ref-table}
#| echo: false
#| message: false
#| results: asis
BibOptions(check.entries = FALSE)
bib <- ReadBib("../references.bib", check = FALSE)

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

fmt_author <- function(e) {
  a <- e$author
  if (is.null(a) || length(a) == 0) return("")
  parts <- character(length(a))
  for (j in seq_along(a)) {
    fam <- paste(a[[j]]$family, collapse = " ")
    giv <- a[[j]]$given
    giv <- giv[nchar(giv) > 0]
    ini <- if (length(giv) > 0) paste0(substr(giv, 1, 1), ".") else character(0)
    parts[j] <- if (length(ini) > 0) paste0(fam, ", ", paste(ini, collapse = " ")) else fam
  }
  paste(parts, collapse = "; ")
}

get_label <- function(e) sub(".*Data file label: (\\w+).*", "\\1", e$note %||% "")

label_order <- c(LETTERS, "AA")
rows <- lapply(seq_along(bib), function(i) {
  e <- bib[i]
  list(label = get_label(e), author = fmt_author(e), citation = e$citation_md %||% "")
})
rows <- rows[order(match(vapply(rows, `[[`, character(1), "label"), label_order))]

cat("| Ref | Author(s) | Citation or website |\n")
cat("|-----|-----------|---------------------|\n")
for (r in rows) cat(sprintf("| %s | %s | %s |\n", r$label, r$author, r$citation))
```
