---
title: "Looking up an unexplained mass difference"
subtitle: "Problem: I have a delta between two peaks and no formula yet. What could it be?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    fig-format: svg
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: svg
vignette: >
  %\VignetteIndexEntry{Looking up an unexplained mass difference}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(commonMZ)
library(ggplot2)

COL <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
        S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
```

[*Reading isotopic fine structure*](isotope-fine-structure-explained.html) and its
companion tool both start from a candidate formula. Most of the time you don't have
one yet — you have two peaks in a spectrum and a delta between them, and the actual
question is "what family of thing produces a gap this size?" An in-source fragment
loses a neutral (water, CO₂, a whole side chain). A homologous series steps by a
repeating unit (CH₂, C₂H₄O, a whole PEG monomer). An adduct swaps one attached ion
for another (Na⁺ for K⁺, formate for acetate). All three are catalogued in commonMZ
already — `adducts_fragments` and `repeating_units_pos`/`repeating_units_neg` — and
`mz_diff_table()` merges them into one searchable reference:

```{r diff-table-build}
diffs <- mz_diff_table("both")
nrow(diffs)
```

Isotope spacings are deliberately not in this table — they depend on which formula
produced the peaks, so they belong to
[the fine-structure tool](isotope-fine-structure-tool.html) instead.

## The map: where do the known differences sit?

Before searching for one specific value, it helps to see the whole landscape at
once — whether your observed delta lands inside a crowded region (several
candidate explanations) or an empty one (nothing catalogued matches, which is
itself useful information). Hover any point for its origin:

```{r diff-map}
#| fig-cap: "Every commonMZ adduct/fragment and repeating-unit mass difference, positioned on the same Da axis your spectrum uses. Hover a point for its origin."
p <- ggplot(diffs, aes(mz_diff, category, colour = category,
                       text = sprintf("%.4f Da — %s\n%s", mz_diff, mode, origin))) +
  geom_point(size = 2, alpha = .7, position = position_jitter(height = 0.15, seed = 1)) +
  scale_colour_manual(values = c(`adduct/fragment` = COL[["C"]], `repeating unit` = COL[["S"]]),
                      guide = "none") +
  labs(x = "m/z difference (Da)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())

plotly::ggplotly(p, tooltip = "text")
```

## The calculator: search by a difference and a ppm tolerance

This is `mz_diff_lookup()`'s job for a single value called from R. Tolerance is in
ppm of the difference by default — the usual way instrument accuracy is quoted —
though note that is *not* the same as ppm of either peak's own m/z; see the
function's documentation (`?mz_diff_lookup`) for why that distinction matters and
when to widen the tolerance or pass an absolute `unit = "Da"` window instead.

```{r lookup-example}
knitr::kable(mz_diff_lookup(18.0106, tol = 100))   # water, 100 ppm of the delta
```

And a delta that is genuinely ambiguous without more context — several catalogued
differences sit within a few ppm of each other, so the delta alone does not
decide; you still need to ask whether the sample plausibly contains an alkane
chain, an acrylamide adduct, or neither:

```{r lookup-example-2}
knitr::kable(mz_diff_lookup(14.0157, tol = 50))
```

## Working interactively

For browsing without writing R code, type a difference and a ppm tolerance and
click *Search* — this is the electronic version of the `first_iso_diff`-style
spreadsheets this tool is descended from. The table's own *Search:* box (top
right) still works as usual for browsing by category, mode or origin text.

```{r diff-interactive}
library(DT)

## the exact DataTables core version DT bundles, for the CDN fallback below --
## read from the widget's own dependency metadata rather than hard-coded, so
## it can't drift out of sync with whatever DT version is installed.
mdl_widget_probe <- datatable(data.frame(x = 1))
DT_CORE_VERSION <- Filter(function(d) d$name == "dt-core", mdl_widget_probe$dependencies)[[1]]$version

htmltools::tagList(
  htmltools::tags$div(
    style = "display:flex; gap:1em; align-items:end; margin-bottom:0.75em; flex-wrap:wrap;",
    htmltools::tags$div(
      htmltools::tags$label("m/z difference (Da)", `for` = "mdl_mz"),
      htmltools::tags$br(),
      htmltools::tags$input(id = "mdl_mz", type = "number", step = "any",
                            class = "form-control", style = "width:12em;")
    ),
    htmltools::tags$div(
      htmltools::tags$label("tolerance (ppm)", `for` = "mdl_ppm"),
      htmltools::tags$br(),
      htmltools::tags$input(id = "mdl_ppm", type = "number", step = "any", value = "50",
                            class = "form-control", style = "width:10em;")
    ),
    htmltools::tags$button("Search", id = "mdl_go", type = "button", class = "btn btn-primary"),
    htmltools::tags$button("Clear", id = "mdl_clear", type = "button", class = "btn btn-outline-secondary")
  ),
  datatable(diffs, rownames = FALSE, elementId = "mdl_table",
           colnames = c("mz_diff (Da)", "category", "mode", "origin", "reference"),
           options = list(pageLength = 8, order = list(list(0, "asc")))),
  htmltools::tags$script(htmltools::HTML(sprintf("
    // Some documentation-site rebuilds of this page have been observed to drop
    // the bundled DataTables core script while keeping its CSS (a
    // pkgdown/htmlwidgets packaging issue, not a data issue) -- which would
    // otherwise also break DT's OWN widget initialisation, not just the code
    // below (both happen on document-ready). document.write() is deprecated
    // for most uses, but this is its one still-legitimate case: called
    // synchronously from a plain, non-deferred inline script WHILE THE PAGE
    // IS STILL PARSING, it inserts the <script> tag directly into the parse
    // stream, so the browser fetches and runs it before parsing continues --
    // i.e. before DOMContentLoaded, before jQuery's ready queue, and before
    // DT's own widget render. That ordering guarantee is why this, and not
    // an appendChild()'d script, is used here.
    if (typeof $ === 'undefined' || typeof $.fn.dataTable === 'undefined') {
      document.write('<script src=\"https://cdn.datatables.net/%s/js/jquery.dataTables.min.js\"><\\/script>');
    }
    $(document).ready(function() {
      // DT's elementId lands on the WRAPPER DIV, not the <table> inside it --
      // grab the real table (and its live DataTables API), and compare DOM
      // nodes rather than ids in the search predicate below.
      var $tbl = $('#mdl_table').find('table').first();
      if (!$tbl.length) return;
      var dt = $tbl.DataTable();
      var target = null, tolDa = null;

      $.fn.dataTable.ext.search.push(function(settings, data, dataIndex) {
        if (settings.nTable !== $tbl.get(0)) return true;
        if (target === null) return true;
        var val = parseFloat(data[0]);
        return !isNaN(val) && Math.abs(val - target) <= tolDa;
      });
      $('#mdl_go').on('click', function() {
        var mz  = parseFloat(document.getElementById('mdl_mz').value);
        var ppm = parseFloat(document.getElementById('mdl_ppm').value);
        if (isNaN(mz) || isNaN(ppm)) { target = null; }
        else { target = mz; tolDa = Math.abs(mz) * ppm / 1e6; }
        dt.draw();
      });
      $('#mdl_clear').on('click', function() {
        document.getElementById('mdl_mz').value = '';
        target = null;
        dt.draw();
      });
    });
  ", DT_CORE_VERSION)))
)
```

The search only re-filters when you click *Search* (not on every keystroke), so
typing a value never triggers a redraw mid-edit -- click once you've entered both
numbers, or press *Clear* to go back to the full table.

## Checklist

1. **Search first, don't guess one hypothesis at a time.** `mz_diff_lookup()` (or
   the interactive table above) checks your delta against every known
   adduct/fragment/repeating-unit difference in commonMZ at once.
2. **A delta rarely has only one explanation.** When several catalogued
   differences land within tolerance of each other, use what you know about the
   sample — the matrix, the method, what's chemically plausible — to break the
   tie, the same way the mobile-phase composition was what actually resolved the
   ambiguous formulas in `contaminants_+.tsv`'s metal-complex entries.
3. **Mind the tolerance convention.** ppm of a small delta is a much tighter
   window than ppm of a large one; if nothing matches, try a wider tolerance or an
   absolute Da window before concluding the delta is uncatalogued.
4. **If you do have (or can guess) a formula**, switch to
   [*Simulating isotope fine structure for a formula*](isotope-fine-structure-tool.html) —
   an isotope satellite is a different kind of delta than this table covers.
