---
title: "Looking up an unexplained mass difference"
subtitle: "Problem: I have a delta between two peaks and no formula yet. What could it be?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
    fig-format: svg

execute:
  warning: false
  message: false
  echo: true
knitr:
  opts_chunk:
    dev: svglite
vignette: >
  %\VignetteIndexEntry{Looking up an unexplained mass difference}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(commonMZ)
library(ggplot2)
library(dplyr)
library(DT)
library(RefManageR)

COL <- c(C = "#2a78d6", H = "#eb6834", N = "#1baf7a", O = "#eda100",
        S = "#e87ba4", Cl = "#008300", combo = "#4a3aa7")
```

```{r fix-dt-core}
#| echo: false
htmltools::includeScript(
  system.file("htmlwidgets/lib/datatables/js/jquery.dataTables.min.js", package = "DT")
)
```

You have two peaks in a spectrum and a delta between them, and the question is "what does this gap correspond to?".

* It could be a fragment, usually a neutral loss, (water, CO₂, a whole side chain).
* It could also be an adduct (Na⁺ for K⁺, formate for acetate).
* Fragments sometimes appear in series of repeating units; e.g. sequential loss of CH₂ in an alkane chain.

All three are catalogued in commonMZ and accessible as individual datasets:

- [`adducts_fragments`](https://github.com/stanstrup/commonMZ/blob/master/inst/extdata/adducts_fragments.tsv) — adduct and neutral-loss mass differences, mode-agnostic
- [`repeating_units_pos`](https://github.com/stanstrup/commonMZ/blob/master/inst/extdata/repeating_units_%2B.tsv) — homologous-series steps in positive mode
- [`repeating_units_neg`](https://github.com/stanstrup/commonMZ/blob/master/inst/extdata/repeating_units_-.tsv) — homologous-series steps in negative mode

`mz_diff_table()` merges them into one searchable reference, optionally filtered by mode:

```{r diff-table-build}
#| echo: true
#| code-fold: false

commonMZ::adducts_fragments     # adducts and neutral losses only
commonMZ::repeating_units_pos   # repeating units, positive mode only

diffs <- mz_diff_table("both")  # all three merged; use "pos" or "neg" to filter
```


## The calculator: search by a difference and a ppm tolerance

This is `mz_diff_lookup()`'s job for a single value called from R.

An instrument's mass accuracy is quoted as ppm of a *measured m/z*, so the ppm
error of a mass *difference* only means something when you say which m/z it is
relative to. That is what `ref_mz` is for: pass the m/z of the parent ion the two
peaks were measured at, and both the ppm tolerance and the returned `error_ppm`
column are expressed relative to it. In the examples below `ref_mz = 300` stands
in for a typical parent ion, so `error_ppm` reads as "how far off would this
assignment be for a compound around m/z 300". Without `ref_mz` the ppm is taken of
the difference itself, which inflates it badly for small deltas — a 0.002 Da error
on a 1 Da difference is 2000 ppm of the delta but only 6.7 ppm at m/z 300. See
`?mz_diff_lookup` for the full argument.

```{r lookup-example}
#| echo: true
#| code-fold: false
#| eval: false
mz_diff_lookup(18.0106, tol = 100, ref_mz = 300) %>%   # water, 100 ppm at m/z 300
  mutate(error_Da = signif(error_Da, 2), error_ppm = round(error_ppm, 1))
```

```{r lookup-example-dt}
#| echo: false
mz_diff_lookup(18.0106, tol = 100, ref_mz = 300) %>%
  mutate(error_Da = signif(error_Da, 2), error_ppm = round(error_ppm, 1)) %>%
  DT::datatable(rownames = FALSE, options = list(pageLength = 10, dom = "t"))
```

The 100 ppm window is 0.03 Da at m/z 300, wide enough to also catch the F ↔ H
halogen exchange two hundredths of a Da away — at −67 ppm, clearly distinguishable
from water's −0.1 ppm.

And a genuinely ambiguous case: a 44 Da delta returns three completely different
explanations — CO₂ neutral loss (decarboxylation), a
double-sodium salt adduct, and a PEG repeat-unit
step (polymer contamination from the LC system). A flat Da window is used here
because the three entries span 62 mDa, more than even a 50 ppm window at m/z 300
(15 mDa) would cover; `ref_mz = 300` is still passed so the reported `error_ppm`
stays on the same realistic scale. The key point is that mass alone cannot decide
between them:

```{r lookup-example-2}
#| echo: true
#| code-fold: false
#| eval: false
mz_diff_lookup(44, tol = 0.05, unit = "Da", ref_mz = 300) %>%
  mutate(error_Da = signif(error_Da, 2), error_ppm = round(error_ppm, 1))
```

```{r lookup-example-2-dt}
#| echo: false
mz_diff_lookup(44, tol = 0.05, unit = "Da", ref_mz = 300) %>%
  mutate(error_Da = signif(error_Da, 2), error_ppm = round(error_ppm, 1)) %>%
  DT::datatable(rownames = FALSE, options = list(pageLength = 10, dom = "t"))
```

## Working interactively

Here is a little interactive table to search for these differences:

```{r diff-interactive}
#| echo: false
htmltools::tagList(
  htmltools::tags$div(
    style = "display:flex; gap:1em; align-items:end; margin-bottom:0.75em; flex-wrap:wrap;",
    htmltools::tags$div(
      htmltools::tags$label("m/z difference (Da)", `for` = "mdl_mz"),
      htmltools::tags$br(),
      htmltools::tags$input(id = "mdl_mz", type = "number", step = "any",
                            class = "form-control", style = "width:12em;")
    ),
    htmltools::tags$div(
      htmltools::tags$label("tolerance (ppm)", `for` = "mdl_ppm"),
      htmltools::tags$br(),
      htmltools::tags$input(id = "mdl_ppm", type = "number", step = "any", value = "50",
                            class = "form-control", style = "width:10em;")
    ),
    htmltools::tags$div(
      htmltools::tags$label("reference m/z (Da)", `for` = "mdl_refmz"),
      htmltools::tags$br(),
      htmltools::tags$input(id = "mdl_refmz", type = "number", step = "any", value = "300",
                            class = "form-control", style = "width:10em;")
    ),
    htmltools::tags$button("Search", id = "mdl_go", type = "button", class = "btn btn-primary"),
    htmltools::tags$button("Clear", id = "mdl_clear", type = "button", class = "btn btn-outline-secondary")
  ),
  datatable(diffs %>% mutate(ppm_error = NA_real_),
            rownames = FALSE, elementId = "mdl_table",
            colnames = c("mz_diff (Da)", "category", "mode", "origin", "reference", "ppm error"),
            options = list(pageLength = 8, order = list(list(0, "asc")))),
  htmltools::includeScript("mass-diff-search.js")
)
```

The search only re-filters when you click *Search* (not on every keystroke), so
typing a value never triggers a redraw mid-edit; click once you've entered both
numbers, or press *Clear* to go back to the full table.


## Glossary

### Ion types

| Notation | Meaning |
|---|---|
| `f+` | fragment ion |
| `[f+H]+` | protonated fragment ion (e.g. in-source fragmentation) |
| `[M+H]+` | protonated molecular ion (pseudomolecular ion) |
| `[M+Na]+` | sodiated molecular ion |
| `[M+K]+` | potassiated molecular ion |
| `[2M+H]+`, `[3M+H]+` | protonated dimer, trimer, etc. |
| `[AnBm+H]+` | protonated ion of a complex with *n* A and *m* B subunits |

### Abbreviations used in the *origin* column

| Abbreviation | Meaning |
|---|---|
| 4-HCCA | α-cyano-4-hydroxycinnamic acid — common MALDI matrix |
| 2,5-DHB | 2,5-dihydroxybenzoic acid — common MALDI matrix |
| MeCN, ACN | acetonitrile (solvent) |
| MeOH | methanol (solvent) |
| MeNO₂ | nitromethane (solvent) |
| HABA | 2-(4-hydroxyphenylazo)benzoic acid — MALDI matrix |
| SA | sinapic / sinapinic acid — common MALDI matrix |
| PEG | polyethylene glycol; repeat unit –[O–CH₂–CH₂]–, 44 Da |
| PPG | polypropylene glycol; repeat unit –[O–C(CH₃)H–CH₂]–, 58 Da |
| XaaCcamXaa | carbamidomethylated cysteine residue (+57 Da) |
| XaaMoxXaa | singly oxidised methionine residue (+16 Da) |

### References for the *reference* column

```{r ref-table}
#| echo: false
#| message: false
#| results: asis
BibOptions(check.entries = FALSE)
bib <- ReadBib("references.bib", check = FALSE)

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

fmt_author <- function(e) {
  a <- e$author
  if (is.null(a) || length(a) == 0) return("")
  parts <- character(length(a))
  for (j in seq_along(a)) {
    fam <- paste(a[[j]]$family, collapse = " ")
    giv <- a[[j]]$given
    giv <- giv[nchar(giv) > 0]
    ini <- if (length(giv) > 0) paste0(substr(giv, 1, 1), ".") else character(0)
    parts[j] <- if (length(ini) > 0) paste0(fam, ", ", paste(ini, collapse = " ")) else fam
  }
  paste(parts, collapse = "; ")
}

get_label <- function(e) sub(".*Data file label: (\\w+).*", "\\1", e$note %||% "")

get_url <- function(e) {
  u <- e$url
  if (!is.null(u) && length(u) > 0 && nchar(trimws(u)) > 0) return(trimws(u))
  d <- e$doi
  if (!is.null(d) && length(d) > 0 && nchar(trimws(d)) > 0)
    return(paste0("https://doi.org/", trimws(d)))
  NULL
}

label_order <- c(LETTERS, "AA")
rows <- lapply(seq_along(bib), function(i) {
  e   <- bib[i]
  cit <- e$citation_md %||% ""
  url <- get_url(e)
  if (!is.null(url)) cit <- sprintf("[%s](%s)", cit, url)
  list(label = get_label(e), author = fmt_author(e), citation = cit)
})
rows <- rows[order(match(vapply(rows, `[[`, character(1), "label"), label_order))]

cat("| Ref | Author(s) | Citation or website |\n")
cat("|-----|-----------|---------------------|\n")
for (r in rows) cat(sprintf("| %s | %s | %s |\n", r$label, r$author, r$citation))
```
